---
title: "PCAngsd"
author: "Mike Connelly"
date: "2023-05-10"
output: html_document
---
## Setup and packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
options(stringsAsFactors = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```
```{r packages}
library("tidyverse")
library("ggrepel")
#
source("./R/pocillopora_etp_gksim_functions.R")
```

## Import sample metadata
```{r import_sample_metadata}
source("./R/sample_import.R")
samples <- all_samples
```
```{r select_samples}
samples <- filter(all_samples, ANALYSIS ==T)
# samples <- filter(all_samples, `PANAMA_MANUSCRIPT` == T)
```
Note: Remove clones from the dataset using IBS analysis before running PCAngsd

## Perform PCA of all samples
```{r pca}
# pop<-read.table("Demo/Data/Demo1pop.info")
C <- as.matrix(read.table("./outputs/angsd/all_analysis.cov"))
e <- eigen(C)
```
```{r pca_test_plot}
plot(e$vectors[,1:2],xlab="PC1",ylab="PC2", main="individual allele frequency")
#
plot(e$vectors[,2:3],xlab="PC2",ylab="PC3", main="individual allele frequency")
# 
```
```{r combine_pca_metadata}
# combine pc coordinates with sample metadata
pcs <- as.data.frame(e$vectors) %>% cbind(samples)
# store variable with % variance explained by each PC
pcvar <- round(e$values, 2)
#
```
### Plot PCA of all-samples
```{r pca_test_ggplot}
title <- expression(paste("PCA of East Pacific ", italic("Pocillopora"), " samples"))
# test ggplot 
pcs %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Region`), shape = 21, size = 2, alpha = 0.75) +
  scale_fill_manual(values = region_colors) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar[2]/pcvar[1]) + 
  xlab(paste0( "PC1 (", pcvar[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar[2], "%)")) +
  ggtitle(title)
# 
pcs %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `mtorf_type`), shape = 21, size = 2, alpha = 0.75) +
  # scale_fill_manual(values = region_colors) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar[2]/pcvar[1]) +
  xlab(paste0( "PC1 (", pcvar[1], "%)")) +
  ylab(paste0( "PC2 (", pcvar[2], "%)")) +
  ggtitle(title)
# 
pcs %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Species`), shape = 21, size = 2, alpha = 0.75) +
  scale_fill_manual(values = spp_colors) +
  # scale_shape_manual(values = c(21:25)) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar[2]/pcvar[1]) + 
  xlab(paste0( "PC1 (", pcvar[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar[2], "%)")) +
  ggtitle(title)
# 
```

```{r figure 1: pca_1-2}
# use custom function to plot PCA colored by Region:Site
pdf("./outputs/figures/pca12.pdf", width = 8, height = 6)
gg_pcangsd(e, samples, pc=c(1,2)) +
  ggtitle(title)
  #
gg_pcangsd(e, samples, pc=c(1,2)) +
  facet_grid(Region ~ .) +
  ggtitle(title)
dev.off()
```
```{r pca_2-3}
pdf("./outputs/figures/pca23.pdf", width = 8, height = 6)
gg_pcangsd(e, samples, pc=c(2,3)) +
  ggtitle(title)
  #
dev.off()
```

## Perform PCA of no-clones samples
```{r pca}
# pop<-read.table("Demo/Data/Demo1pop.info")
C <- as.matrix(read.table("./outputs/angsd/full_noclones.cov"))
e <- eigen(C)
```
```{r pca_test_plot}
plot(e$vectors[,c(1,2)],xlab="PC1",ylab="PC2", main="individual allele frequency")
#
```

```{r bamfile_sample_order_check}
# reading list of bam files = order of samples in IBS matrix
bams <- read.table("/Users/mikeconnelly/computing/projects/etp_pocillopora_gskim/outputs/angsd/full_noclones_bamfile.txt",header=F)[,1]
#
bam_ids <- sub("/scratch/nmnh_corals/connellym/projects/etp_pocillopora_gskim/outputs/alignments/", "", bams, perl=T)
bam_ids <- sub(".sorted.md.rg.bam", "", bam_ids, perl=T)
bam_ids <- sub("_S[[:digit:]]+$", "", bam_ids)
#
bam_ids
```
```{r select_samples}
# aligning sample metadata with bams list, then check again
samples_noclones <- samples[match(bam_ids, samples$`Sample ID`),] 
# do all the sample ids match the bam file ids in the correct order?
all(samples_noclones$`Sample ID` == bam_ids) # TRUE
```
```{r combine_pca_metadata}
# combine pc coordinates with sample metadata
pcs <- as.data.frame(e$vectors) %>% cbind(samples_noclones)
# store variable with % variance explained by each PC
pcvar <- round(e$values, 2)
#
```

```{r}
pdf(file="./outputs/figures/sicb/pcangsd_species.pdf", width = 5.5, height = 3.75)
title <- expression(paste("East Pacific ", italic("Pocillopora"), " PCA (4,233,691 SNPs)"))
pca <- pcs %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Species`), shape = 21, size = 3, alpha = 0.75, show.legend = F) +
  scale_fill_manual(values = spp_colors) +
  # facet_grid(Region ~ .) +
  # coord_fixed(pcvar[2]/pcvar[1]) +
  xlab(paste0( "PC1 (", pcvar[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar[2], "%)")) +
  theme(plot.title = element_text(size = 16, hjust = -0.2),
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)) +
  ggtitle(title)
dev.off()
```
```{r}
png(file="./outputs/figures/sicb/pcangsd_species.png", width = 5.5, height = 3.75, units = "in",  res = 300, type = "cairo-png")
pca
dev.off()
```


### PCA of all samples by species
```{r}
pdf(file="./outputs/figures/etp_spp_pca_color.pdf", width = 4, height = 2.5)
title <- expression(paste("PCA of East Pacific ", italic("Pocillopora"), " spp."))
pcs %>% ggplot(aes(x = V1, y = V2)) +
  geom_jitter(aes(fill = `Species`), shape = 21, size = 4, alpha = 0.6, show.legend = T, width = 0.005, height = 0.005) +
  scale_fill_manual(values = spp_colors) +
  facet_wrap(Region ~ .) +
  coord_fixed(pcvar[3]/pcvar[2]) +
  xlab(paste0( "PC1 (", pcvar[2], "%)")) + 
  ylab(paste0( "PC2 (", pcvar[3], "%)")) +
  theme(plot.title = element_text(size = 16, hjust = 1.4),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 11),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12, face = "italic")) +
  ggtitle(title)
dev.off()
```


## Perform separate PCAs for different species

### Plot PCAs for *P. grandis* - no clones
```{r pca_c1}
C1 <- as.matrix(read.table("./outputs/angsd/pgrandis_noclones.cov"))
e1 <- eigen(C1)
```
```{r pca_test_plot}
plot(e1$vectors[,1:2],xlab="PC1",ylab="PC2", main="individual allele frequency")
#
plot(e1$vectors[,2:3],xlab="PC2",ylab="PC3", main="individual allele frequency")
# 
```
```{r bamfile_sample_order_check}
# reading list of bam files = order of samples in IBS matrix
bams <- read.table("/Users/mikeconnelly/computing/projects/etp_pocillopora_gskim/outputs/angsd/pgrandis_noclones_bamfile.txt",header=F)[,1]
#
bam_ids <- sub("/scratch/nmnh_corals/connellym/projects/etp_pocillopora_gskim/outputs/alignments/", "", bams, perl=T)
bam_ids <- sub(".sorted.md.rg.bam", "", bam_ids, perl=T)
bam_ids <- sub("_S[[:digit:]]+$", "", bam_ids)
#
# bam_ids
```
```{r select_samples}
# aligning sample metadata with bams list, then check again
samples_pgrandis <- samples[match(bam_ids, samples$`Sample ID`),] 
# do all the sample ids match the bam file ids in the correct order?
all(samples_pgrandis$`Sample ID` == bam_ids) # TRUE
```
```{r combine_pca_metadata}
# combine pc coordinates with sample metadata
pcs_c1 <- as.data.frame(e1$vectors) %>% cbind(samples_pgrandis)
# store variable with % variance explained by each PC
pcvar_c1 <- round(e1$values, 2)
#
```
```{r pca_metadata_factors}
pcs_c1$Region <- factor(pcs_c1$Region, levels = c("Panama", "Chiriqui", "Golfo Dulce", "Bahia Culebra", "Gorgona", "Coco", "Galapagos Central", "Galapagos Northern", "Clipperton"), ordered = T)
```

```{r pca_ggplot}
# test ggplot 
pcs_c1 %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Region`), shape = 21, size = 2, alpha = 0.75) +
  scale_fill_manual(values = region_colors) +
  facet_wrap(~ `Region:Site`) +
  coord_fixed(pcvar_c1[2]/pcvar_c1[1]) + 
  theme(legend.position = "right") +
  xlab(paste0( "PC1 (", pcvar_c1[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar_c1[2], "%)"))
# 
# test ggplot 
pcs_c1 %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Region:Site`), shape = 21, size = 2, alpha = 0.75) +
  # scale_fill_manual(values = site_colors) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar_c1[2]/pcvar_c1[1]) + 
  theme(legend.position = "right") +
  xlab(paste0( "PC1 (", pcvar_c1[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar_c1[2], "%)"))
# 
```
```{r pgrandis_region_pca}
pdf("./outputs/figures/pgrandis_region_pca.pdf", height = 4.5, width = 6)
title <- expression(paste("PCA of ETP ", italic("P. grandis"), " samples (n = 173)"))
#
pcs_c1 %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Region`), shape = 21, size = 2, alpha = 0.75) +
  scale_fill_manual(values = region_colors) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar_c1[2]/pcvar_c1[1]) + 
  theme(legend.position = "right") +
  xlab(paste0( "PC1 (", pcvar_c1[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar_c1[2], "%)")) +
  ggtitle(title)
# 
dev.off()
```

### Plot PCAs for P. grandis - Panama
```{r pca_c1}
pgrandis <- read_csv("./data/pgrandis_panama_samples.txt", col_names = "Sample ID") %>% arrange(`Sample ID`)
pgrandis_samples <- pgrandis %>% left_join(samples)
#
C1 <- as.matrix(read.table("./outputs/angsd/pgrandis_panama.cov"))
e1 <- eigen(C1)
```
```{r pca_test_plot}
plot(e1$vectors[,1:2],xlab="PC1",ylab="PC2", main="individual allele frequency")
#
plot(e1$vectors[,2:3],xlab="PC2",ylab="PC3", main="individual allele frequency")
# 
```
```{r combine_pca_metadata}
# combine pc coordinates with sample metadata
pcs_c1 <- as.data.frame(e1$vectors) %>% cbind(pgrandis_samples)
# store variable with % variance explained by each PC
pcvar_c1 <- round(e1$values, 2)
#
```
```{r pca_ggplot}
# test ggplot 
pdf("~/Desktop/pgrandis_subspecies.pdf", width = 8.5)
pcs_c1 %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Country`), shape = 21, size = 2, alpha = 0.75) +
  geom_text_repel(aes(label = `Sample ID`), size = 1, max.overlaps = 50) +
  # scale_fill_manual(values = region_colors) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar_c1[2]/pcvar_c1[1]) + 
  theme(legend.position = "right") +
  xlab(paste0( "PC1 (", pcvar_c1[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar_c1[2], "%)"))
dev.off()
# 
# test ggplot 
pcs_c1 %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Region:Site`), shape = 21, size = 2, alpha = 0.75) +
  # scale_fill_manual(values = site_colors) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar_c1[2]/pcvar_c1[1]) + 
  theme(legend.position = "right") +
  xlab(paste0( "PC1 (", pcvar_c1[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar_c1[2], "%)"))
# 
```
```{r}
pcs_c1_spp <- pcs_c1 %>% mutate(SubSpecies = ifelse(V2 < -0.05, "C", (ifelse(V1 < -0.05, "B", ifelse(V1 > 0, "A", "D"))))) 
```
```{r}
pdf(file="./outputs/figures/sicb/pcangsd_pgrandis_region.pdf", width = 7.5, height = 5)
title <- expression(paste("East Pacific ", italic("P. grandis"), " population-level PCA (3,273,302 SNPs)"))
# no bone Pcas 
pcs_c1 %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Region`), shape = 21, size = 2, alpha = 0.75, show.legend = T) +
  scale_fill_manual(values = region_colors) +
  # facet_wrap(~ Region) +
  coord_fixed(pcvar_c1[2]/pcvar_c1[1]) + 
  xlab(paste0( "PC1 (", pcvar_c1[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar_c1[2], "%)")) + 
  theme(plot.title = element_text(size = 16, hjust = 0),
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.position = "right", 
        legend.spacing.y = unit(2, "mm"),
        legend.title = element_blank(),
        legend.text = element_text(size = 12)) +
  ggtitle(title)
dev.off()
```
```{r}
png(file="./outputs/figures/sicb/pcangsd_pgrandis_region.png", width = 8, height = 3.75, units = "in", res = 300)
title <- expression(paste(italic("P. grandis"), " population-level PCA (3,273,302 SNPs)"))
# no bone Pcas 
pcs_c1 %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Region`), shape = 21, size = 2, alpha = 0.75, show.legend = T) +
  scale_fill_manual(values = region_colors) +
  # facet_wrap(~ Region) +
  coord_fixed(pcvar_c1[2]/pcvar_c1[1]) + 
  xlab(paste0( "PC1 (", pcvar_c1[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar_c1[2], "%)")) + 
  theme(plot.title = element_text(size = 16, hjust = 0.1),
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.position = "right", 
        legend.spacing.y = unit(2, "mm"),
        legend.title = element_blank(),
        legend.text = element_text(size = 12)) +
  ggtitle(title)
dev.off()
```

### Plot PCAs for *P. verrucosa* (3a + 3b + 3d) - no clones
```{r pca_c3}
C <- as.matrix(read.table("./outputs/angsd/pverrucosa_full_noclones.cov"))
e <- eigen(C)
```
```{r pca_test_plot}
plot(e$vectors[,1:2],xlab="PC1",ylab="PC2", main="individual allele frequency")
#
plot(e$vectors[,2:3],xlab="PC2",ylab="PC3", main="individual allele frequency")
# 
```
```{r bamfile_sample_order_check}
# reading list of bam files = order of samples in IBS matrix
bams <- read.table("/Users/mikeconnelly/computing/projects/etp_pocillopora_gskim/outputs/angsd/pverrucosa_full_noclones_bamfile.txt",header=F)[,1]
#
bam_ids <- sub("/scratch/nmnh_corals/connellym/projects/etp_pocillopora_gskim/outputs/alignments/", "", bams, perl=T)
bam_ids <- sub(".sorted.md.rg.bam", "", bam_ids, perl=T)
bam_ids <- sub("_S[[:digit:]]+$", "", bam_ids)
#
# bam_ids
```
```{r select_samples}
# aligning sample metadata with bams list, then check again
samples_pverrucosa <- samples[match(bam_ids, samples$`Sample ID`),] 
# do all the sample ids match the bam file ids in the correct order?
all(samples_pverrucosa$`Sample ID` == bam_ids) # TRUE
```
```{r combine_pca_metadata}
# combine pc coordinates with sample metadata
pcs <- as.data.frame(e$vectors)# %>% cbind(pverrucosa) %>% left_join(samples)
# store variable with % variance explained by each PC
pcvar <- round(e$values, 2)
#
```
```{r combine_pca_metadata}
# combine pc coordinates with sample metadata
pcs <- as.data.frame(e$vectors) %>% cbind(samples_pverrucosa)
# store variable with % variance explained by each PC
pcvar <- round(e$values, 2)
#
```
```{r pca_cluster_ggplot}
# test ggplot 
pcs %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Region`), shape = 21, size = 2, alpha = 0.75) +
  scale_fill_manual(values = region_colors) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar[2]/pcvar[1]) +
  xlab(paste0( "PC1 (", pcvar[1], "%)")) +
  ylab(paste0( "PC2 (", pcvar[2], "%)"))
# 
```
```{r pverrucosa_region_pca}
pdf("./outputs/figures/pverrucosa_region_pca.pdf", width = 8, height = 3.75)
title <- expression(paste(italic("P. verrucosa"), " population-level PCA (42 samples, 13,224,661 SNPs)"))
#
pcs %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Region`), shape = 21, size = 2, alpha = 0.75) +
  scale_fill_manual(values = region_colors) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar[2]/pcvar[1]) + 
  theme(legend.position = "right") +
  xlab(paste0( "PC1 (", pcvar[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar[2], "%)")) +
  theme(plot.title = element_text(size = 16, hjust = 0.1),
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.position = "right", 
        legend.spacing.y = unit(2, "mm"),
        legend.title = element_blank(),
        legend.text = element_text(size = 12)) +
  ggtitle(title)
# 
dev.off()
```