---
title: "PHYLUCE_phylogenetic_analysis"
author: "Mike Connelly"
date: "2023-05-24"
output: html_document
---
## Setup and packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
options(stringsAsFactors = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```
```{r packages}
library("tidyverse")
library("treeio")
library("ggtree")
library("phytools")
# 
source("./R/pocillopora_etp_gksim_functions.R")
```

## Import sample metadata
```{r import_sample_metadata}
source("./R/sample_import.R")
samples <- all_samples
```
```{r reference_sample_metadata}
# Oury contemporary samples
oury_samples <- read_csv("./data/top5_oury_samples.csv") %>% 
  mutate(tip_label = str_c(GSH, Sample_Name, sep = " ")) # NCBI_SSR,

# Reference genomes

```
```{r outgroup_metadata}
# outgroups <- read_csv("./data/outgroups.csv")
# samples_out <- rbind(samples, outgroups)
```

## IQ-TREE w/ PHYLUCE alignments
### PociGSkim NGSAdmix K=5 top 5 reps plus Clip_16 and Coco_614 (26) + Oury GSH top 5 reps (103) + reference genome samples (8) + outgroups (1) (n = ### samples)
#### 75% matrix
 136 taxa, 1,263 UCE loci, 613,088 bp, 166,712  informative sites
```{r import_tree}
tree <- read.newick("./outputs/iqtree/iqtree_representative_5X_phyluce_iqtree_75.treefile")

# midpoint root
tree <- midpoint.root(tree)

# drop outgroup tips (easier to view node labels)
tree <- ape::drop.tip(tree, tip = "Stylophora_pistillata_ANT17_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052307_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052309_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052305_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052306_contigs")

# set root-edge
tree$root.edge <- 1
# check topology
ggtree(tree)
```
```{r adjust_tip_labels}
# adjust tip labels
# remove contigs suffix 
tips <- gsub("_contigs$", "", tree$tip.label)
# remove sample number suffix 
tips <- gsub("_S[[:digit:]]*$", "", tips)

# Reference genomes: tip names
tips <- gsub("Stylophora_", "S. ", tips)
tips <- gsub("_ANT17", " ANT17", tips)
tips <- gsub("pocillopora_", "P. ", tips)
tips <- gsub("_G", " G", tips)
tips <- gsub("_SAMN", " SAMN", tips)

# Oury samples:
oury_tip_order <- tips[tips %in% oury_samples$NCBI_SSR]

tips[tips %in% oury_samples$NCBI_SSR] <- oury_samples[match(oury_tip_order, oury_samples$NCBI_SSR),]$tip_label

# need to change reference genome labels to include GenBank accessions
# Stephens et al. 2022 genomes do not have GenBank/RefSeq accessions, using BioSample numbers
# SAMN21845732 - P. meandrina
# SAMN22898959 - P. acuta
```
```{r check_tree_tips}
# check tips
tips
# assign new tip labels to tree
tree$tip.label <- tips

# basic viz
gg_tree(tree)
```
```{r tip_metadata}
# create dataframe holding metadata 
tips_metadata <- data.frame(tips)
tips_metadata$class <- ifelse(grepl("Glo", tips), "Outgroup", ifelse(grepl("P\\.|S\\.", tips), "Genome", ifelse(grepl("GSH", tips), "Oury", "Connelly")))
# red for USNM, dark blue for genomes, dark green for Oury, black for Connelly
```
```{r node_labels}
# node labels with bootstrap support
node.labels <- tree$node.label
# preserve original labels
node.labels.original <- node.labels
node.labels <- node.labels.original

# remove root label
node.labels[node.labels == "Root"] <- ""
# remove 100% supported node labels
# node.labels[node.labels == "100/100"] <- ""
# define high-support bootstrap values - >90%?
node.labels[grepl("^100", node.labels)] <- ""
node.labels[grepl("^9", node.labels)] <- ""
# mark low-support bootstrap values with asterisks - <80%?
# node.labels[grepl("^[8]", node.labels)] <- ""
# node.labels[grepl("^[5-7]", node.labels)] <- "<80"
# node.labels[grepl("^[0-4]", node.labels)] <- "<50"

# check to see if formatting is correct
node.labels
#
tree$node.label <- node.labels
```
```{r full_ggtree_viz}
pdf("./outputs/figures/iqtree_representative_75pct_tree.pdf", width = 6.5, height = 8.5)
#
tree_title <- expression(paste("IQ-TREE Maximum Likelihood Tree with reference ", italic("Pocillopora"), " GSH samples"))
loci_info <- c() #c("75% sample occupancy, 1,273 UCE loci, 613,550 bp, 167,666 informative sites")
#

tree %>% ggtree(., right = TRUE) %<+% tips_metadata + 
  # figure out way to highlight clades
  # geom_highlight(node = 69, fill = "orange", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  # geom_highlight(node = 115, fill = "turquoise", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  # geom_highlight(node = 105, fill = "magenta", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  # geom_highlight(node = 62, fill = "purple", type = "gradient", gradient.direction = 'rt', alpha = .6) +
  # geom_highlight(node = 70, fill = "purple", type = "gradient", gradient.direction = 'rt', alpha = .6) +

  # node labels (numeric)
  # geom_text(aes(label=node), hjust=-.3, size = 1) +
  
  # node labels
  geom_nodelab(size = 2, hjust = 0, nudge_y = 0, geom = "shadowtext", color = "grey90", bg.color = "white") +
  
  # tip labels
  geom_tiplab(hjust = 0, aes(color = `class`, size = `class`)) +
  
  # label clades
  # geom_cladelab(node = 41, label = "P. effusa", fontsize = 3, fontface = 3, offset = 0.01, offset.text = 0.0004) +
  # geom_cladelab(node = 47, label = "P. grandis", fontsize = 3, fontface = 3, offset = 0.011, offset.text = 0.0004) +
  # geom_cladelab(node = 60, label = "P. meandrina", fontsize = 3, fontface = 3, offset = 0.014, offset.text = 0.0004) +
  # geom_cladelab(node = 62, label = "P. acuta / damicornis", fontsize = 3, fontface = 3, offset = 0.009, offset.text = 0.0004) +
  # geom_cladelab(node = 70, label = "P. verrucosa", fontsize = 3, fontface = 3, offset = 0.011, offset.text = 0.0004) +

  # setup tree visualization
  geom_rootedge(rootedge = 0.001) +
  geom_treescale(width = 0.01, x = 0, y = -1, linesize = 0.75) +
  ggplot2::xlim(-0.001, 0.06) +
  scale_color_manual(values = c("red", "navyblue", "black")) +
  scale_size_manual(values = c(1.75, 1.75, 1.5, 1.5)) +
  theme(plot.margin = margin(5,2,2,5, unit = "mm"),
        legend.position = "none",
        plot.title = element_blank(),
        plot.subtitle = element_text(size = 10)) +
  ggtitle(tree_title, subtitle = loci_info)
#
dev.off()
```

#### 50% matrix
 136 taxa, 1,513 UCE loci, 717,549 bp, 194,624  informative sites
```{r import_tree}
tree <- read.newick("./outputs/iqtree/iqtree_representative_5X_phyluce_iqtree_50.treefile")

# midpoint root
tree <- midpoint.root(tree)

# drop outgroup tips (easier to view node labels)
tree <- ape::drop.tip(tree, tip = "Stylophora_pistillata_ANT17_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052307_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052309_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052305_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052306_contigs")

# set root-edge
tree$root.edge <- 1
# check topology
ggtree(tree)
```
```{r adjust_tip_labels}
# adjust tip labels
# remove contigs suffix 
tips <- gsub("_contigs$", "", tree$tip.label)
# remove sample number suffix 
tips <- gsub("_S[[:digit:]]*$", "", tips)

# Reference genomes: tip names
tips <- gsub("Stylophora_", "S. ", tips)
tips <- gsub("_ANT17", " ANT17", tips)
tips <- gsub("pocillopora_", "P. ", tips)
tips <- gsub("_G", " G", tips)
tips <- gsub("_SAMN", " SAMN", tips)

# Oury samples:
oury_tip_order <- tips[tips %in% oury_samples$NCBI_SSR]

tips[tips %in% oury_samples$NCBI_SSR] <- oury_samples[match(oury_tip_order, oury_samples$NCBI_SSR),]$tip_label

# need to change reference genome labels to include GenBank accessions
# Stephens et al. 2022 genomes do not have GenBank/RefSeq accessions, using BioSample numbers
# SAMN21845732 - P. meandrina
# SAMN22898959 - P. acuta
```
```{r check_tree_tips}
# check tips
tips
# assign new tip labels to tree
tree$tip.label <- tips

# basic viz
gg_tree(tree)
```
```{r tip_metadata}
# create dataframe holding metadata 
tips_metadata <- data.frame(tips)
tips_metadata$class <- ifelse(grepl("Glo", tips), "Outgroup", ifelse(grepl("P\\.|S\\.", tips), "Genome", ifelse(grepl("GSH", tips), "Oury", "Connelly")))
# red for USNM, dark blue for genomes, dark green for Oury, black for Connelly
```
```{r node_labels}
# node labels with bootstrap support
node.labels <- tree$node.label
# preserve original labels
node.labels.original <- node.labels
node.labels <- node.labels.original

# remove root label
node.labels[node.labels == "Root"] <- ""
# remove 100% supported node labels
# node.labels[node.labels == "100/100"] <- ""
# define high-support bootstrap values - >90%?
node.labels[grepl("^100", node.labels)] <- ""
node.labels[grepl("^9", node.labels)] <- ""
# mark low-support bootstrap values with asterisks - <80%?
# node.labels[grepl("^[8]", node.labels)] <- ""
# node.labels[grepl("^[5-7]", node.labels)] <- "<80"
# node.labels[grepl("^[0-4]", node.labels)] <- "<50"

# check to see if formatting is correct
node.labels
#
tree$node.label <- node.labels
```
```{r full_ggtree_viz}
pdf("./outputs/figures/iqtree_representative_50pct_tree.pdf", width = 6.5, height = 8.5)
#
tree_title <- expression(paste("IQ-TREE Maximum Likelihood Tree with reference ", italic("Pocillopora"), " GSH samples"))
loci_info <- c()# c("50% sample occupancy, 1,519 UCE loci, 713,305 bp, 193,855 informative sites")
#

tree %>% ggtree(., right = TRUE) %<+% tips_metadata + 
  # figure out way to highlight clades
  # geom_highlight(node = 69, fill = "orange", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  # geom_highlight(node = 115, fill = "turquoise", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  # geom_highlight(node = 105, fill = "magenta", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  # geom_highlight(node = 62, fill = "purple", type = "gradient", gradient.direction = 'rt', alpha = .6) +
  # geom_highlight(node = 70, fill = "purple", type = "gradient", gradient.direction = 'rt', alpha = .6) +

  # node labels (numeric)
  # geom_text(aes(label=node), hjust=-.3, size = 1) +
  
  # node labels
  geom_nodelab(size = 2, hjust = 0, nudge_y = 0, geom = "shadowtext", color = "grey90", bg.color = "white") +
  
  # tip labels
  geom_tiplab(hjust = 0, aes(color = `class`, size = `class`)) +
  
  # label clades
  # geom_cladelab(node = 41, label = "P. effusa", fontsize = 3, fontface = 3, offset = 0.01, offset.text = 0.0004) +
  # geom_cladelab(node = 47, label = "P. grandis", fontsize = 3, fontface = 3, offset = 0.011, offset.text = 0.0004) +
  # geom_cladelab(node = 60, label = "P. meandrina", fontsize = 3, fontface = 3, offset = 0.014, offset.text = 0.0004) +
  # geom_cladelab(node = 62, label = "P. acuta / damicornis", fontsize = 3, fontface = 3, offset = 0.009, offset.text = 0.0004) +
  # geom_cladelab(node = 70, label = "P. verrucosa", fontsize = 3, fontface = 3, offset = 0.011, offset.text = 0.0004) +

  # setup tree visualization
  geom_rootedge(rootedge = 0.001) +
  geom_treescale(width = 0.01, x = 0, y = -1, linesize = 0.75) +
  ggplot2::xlim(-0.001, 0.06) +
  scale_color_manual(values = c("red", "navyblue", "black")) +
  scale_size_manual(values = c(1.75, 1.75, 1.5, 1.5)) +
  theme(plot.margin = margin(5,2,2,5, unit = "mm"),
        legend.position = "none",
        plot.title = element_blank(),
        plot.subtitle = element_text(size = 10)) +
  ggtitle(tree_title, subtitle = loci_info)
#
dev.off()
```

### Figure 2
#### 50% matrix
 138 taxa, 1,513 UCE loci, 717,549 bp, 194,624  informative sites
```{r import_tree}
tree <- read.newick("./outputs/iqtree/iqtree_representative_5X_phyluce_iqtree_50.treefile")

# midpoint root
tree <- midpoint.root(tree)

# drop outgroup tips (easier to view node labels)
tree <- ape::drop.tip(tree, tip = "Stylophora_pistillata_ANT17_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052307_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052309_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052305_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052306_contigs")

# set root-edge
tree$root.edge <- 1
# check topology
ggtree(tree)
```
```{r adjust_tip_labels}
# adjust tip labels
# remove contigs suffix 
tips <- gsub("_contigs$", "", tree$tip.label)
# remove sample number suffix 
tips <- gsub("_S[[:digit:]]*$", "", tips)

# Reference genomes: tip names
tips <- gsub("Stylophora_", "S. ", tips)
tips <- gsub("_ANT17", " ANT17", tips)
tips <- gsub("pocillopora_", "P. ", tips)
tips <- gsub("_G", " G", tips)
tips <- gsub("_SAMN", " SAMN", tips)

# Oury samples:
oury_tip_order <- tips[tips %in% oury_samples$NCBI_SSR]

tips[tips %in% oury_samples$NCBI_SSR] <- oury_samples[match(oury_tip_order, oury_samples$NCBI_SSR),]$tip_label

# need to change reference genome labels to include GenBank accessions
# Stephens et al. 2022 genomes do not have GenBank/RefSeq accessions, using BioSample numbers
# SAMN21845732 - P. meandrina
# SAMN22898959 - P. acuta
```
```{r check_tree_tips}
# check tips
tips
# assign new tip labels to tree
tree$tip.label <- tips

# basic viz
gg_tree(tree)
```
```{r tip_metadata}
# create dataframe holding metadata 
tips_metadata <- data.frame(tips)
tips_metadata$class <- ifelse(grepl("Glo", tips), "Outgroup", ifelse(grepl("P\\.|S\\.", tips), "Genome", ifelse(grepl("GSH", tips), "Oury", "Connelly")))
# red for USNM, dark blue for genomes, dark green for Oury, black for Connelly
```
```{r node_labels}
# node labels with bootstrap support
node.labels <- tree$node.label
# preserve original labels
node.labels.original <- node.labels
node.labels <- node.labels.original

# remove root label
node.labels[node.labels == "Root"] <- ""
# remove 100% supported node labels
# node.labels[node.labels == "100/100"] <- ""
# define high-support bootstrap values - >90%?
node.labels[grepl("^100", node.labels)] <- ""
node.labels[grepl("^9", node.labels)] <- ""
# mark low-support bootstrap values with asterisks - <80%?
# node.labels[grepl("^[8]", node.labels)] <- ""
# node.labels[grepl("^[5-7]", node.labels)] <- "<80"
# node.labels[grepl("^[0-4]", node.labels)] <- "<50"

# check to see if formatting is correct
node.labels
#
tree$node.label <- node.labels
```
```{r full_ggtree_viz}
# pdf("./outputs/figures/Fig2_iqtree_rep_50pct_tree.pdf", width = 6.5, height = 8.5)
#
tree_title <- expression(paste("IQ-TREE Maximum Likelihood Tree with reference ", italic("Pocillopora"), " GSH samples"))
loci_info <- c()# c("50% sample occupancy, 1,519 UCE loci, 713,305 bp, 193,855 informative sites")
#

p <- tree %>% ggtree(., right = TRUE) %<+% tips_metadata + 
  
  # highlight clades
  geom_highlight(node = 215, fill = "darkorange", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 145, fill = "turquoise2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 255, fill = "purple3", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 243, fill = "gold2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 50, fill = "gold2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +

  # node labels (numeric)
  geom_text(aes(label=node), hjust=-.3, size = 1) +
  
  # node labels
  geom_nodelab(size = 2, hjust = 0, nudge_y = 0, geom = "shadowtext", color = "grey90", bg.color = "white") +
  
  # tip labels
  geom_tiplab(hjust = 0, aes(color = `class`, size = `class`)) +
  
  # label clades
  geom_cladelab(node = 257, label = "P. effusa", fontsize = 3, fontface = 3, offset = 0.008, offset.text = 0.0004) +
  # 
  geom_cladelab(node = 242, label = "P. meandrina", fontsize = 3, fontface = 3, offset = 0.009, offset.text = 0.0004) +
  # 
  geom_cladelab(node = 218, label = "P. capitata", fontsize = 3, fontface = 3, offset = 0.011, offset.text = 0.0004) +
  # 
  geom_cladelab(node = 149, label = "P. lacera", fontsize = 3, fontface = 3, offset = 0.011, offset.text = 0.0004) +

  # setup tree visualization
  geom_rootedge(rootedge = 0.001) +
  geom_treescale(width = 0.01, x = 0, y = -1, linesize = 0.75) +
  ggplot2::xlim(-0.001, 0.06) +
  scale_color_manual(values = c("red", "navyblue", "black")) +
  scale_size_manual(values = c(1.75, 1.75, 1.5, 1.5)) +
  theme(plot.margin = margin(5,2,2,5, unit = "mm"),
        legend.position = "none",
        plot.title = element_blank(),
        plot.subtitle = element_text(size = 10)) +
  ggtitle(tree_title, subtitle = loci_info)
#

print(p)
# dev.off()
```

```{r cladeI-IV_collapsed_tree}
p <- tree %>% ggtree(., right = TRUE) %<+% tips_metadata

pfinal <- p %>%
  collapse(137, 'min', fill='white', color="black", alpha=.4) %>%
  collapse(170, 'min', fill='white', color="black", alpha=.4) %>%
  collapse(214, 'min', fill='white', color="black", alpha=.4) %>%
  collapse(257, 'min', fill='white', color="black", alpha=.4)
  #

print(pfinal)
```
```{r GSH_collapsed_tree}

p <- tree %>% ggtree(., right = TRUE) %<+% tips_metadata

pfinal <- p %>%
  # clade 2
  collapse(253, 'min', fill='white', color="black", alpha=.4) %>% # GSH 13b
  collapse(249, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(239, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(234, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(230, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(225, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(215, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(163, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(159, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(154, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(138, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(142, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  #
  collapse(197, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(198, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(207, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(203, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(191, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(187, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(183, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(171, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(179, 'min', fill='white', color="black", alpha=.4) %>% # GSH 
  collapse(175, 'min', fill='white', color="black", alpha=.4) # GSH 
  

print(pfinal)

```
```{r custom_collapsed_annotated_tree}
p <- tree %>% ggtree(., right = TRUE) %<+% tips_metadata 

p_collapsed <- p %>%
  
  # # scale clades down
  scaleClade(., node = 171, scale = 0.25) %>%
  # scaleClade(., node = 194, scale = 1) %>%
  # 
  # # scale clades up
  scaleClade(., node = 257, scale = 1.4) %>%
  scaleClade(., node = 242, scale = 1.4) %>%
  scaleClade(., node = 216, scale = 1.4) %>%
  scaleClade(., node = 146, scale = 1.4) %>%
   # scaleClade(., node = 146, scale = 1.4) %>%


  # collapse clades
  collapse(140, 'min', fill='white', color="black", alpha=.4) %>% # GSH15
  collapse(158, 'min', fill='white', color="black", alpha=.4) %>% # GSH13a
  collapse(162, 'min', fill='white', color="black", alpha=.4) %>% # GSH12
  collapse(166, 'min', fill='white', color="black", alpha=.4) %>% # GSH14
  collapse(171, 'min', fill='white', color="black", alpha=.4) %>% # GSH05
  # 
  # collapse(194, 'min', fill='white', color="black", alpha=.4) %>%
  collapse(200, 'min', fill='white', color="black", alpha=.4) %>% # GSH04c
  collapse(201, 'min', fill='white', color="black", alpha=.4) %>% # GSH10
  collapse(206, 'min', fill='white', color="black", alpha=.4) %>% # GSH04b
  collapse(210, 'min', fill='white', color="black", alpha=.4) %>% # GSH04a
  # 
  collapse(237, 'min', fill='white', color="black", alpha=.4) %>% # GSH09a
  collapse(253, 'min', fill='white', color="black", alpha=.4) # GSH13b


p_annotated <- p_collapsed +
  
  # highlight clades
  geom_highlight(node = 218, fill = "darkorange", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 149, fill = "turquoise2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 259, fill = "purple3", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 245, fill = "gold2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +

  # node labels (numeric)
  # geom_text(aes(label=node), hjust=-.3, size = 1) +
  
  # node labels
  geom_nodelab(size = 2, hjust = 0, nudge_y = 0, geom = "shadowtext", color = "grey90", bg.color = "white") +
  
  # tip labels
  geom_tiplab(hjust = 0, aes(color = `class`, size = `class`)) +
  
  # label clades - collapsed GSHs
  # clade 2
  geom_cladelab(node = 237, label = "GSH09a", fontsize = 2, fontface = 1, offset = 0.008, offset.text = 0.0001) +
  geom_cladelab(node = 253, label = "GSH13b", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  # # clade 3
  geom_cladelab(node = 140, label = "GSH15", fontsize = 2, fontface = 1, offset = 0.0065, offset.text = 0.0001) +
  geom_cladelab(node = 158, label = "GSH13a", fontsize = 2, fontface = 1, offset = 0.007, offset.text = 0.0001) +
  geom_cladelab(node = 162, label = "GSH12", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  geom_cladelab(node = 166, label = "GSH14", fontsize = 2, fontface = 1, offset = 0.009, offset.text = 0.0001) +
  # # clade 4
  geom_cladelab(node = 200, label = "GSH04c", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  geom_cladelab(node = 201, label = "GSH10", fontsize = 2, fontface = 1, offset = 0.005, offset.text = 0.0001) +
  geom_cladelab(node = 206, label = "GSH04b", fontsize = 2, fontface = 1, offset = 0.006, offset.text = 0.0001) +
  geom_cladelab(node = 210, label = "GSH04a", fontsize = 2, fontface = 1, offset = 0.007, offset.text = 0.0001) +
  # # 
  geom_cladelab(node = 171, label = "GSH05", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  # #
  
  # label clades - TEP GSHs
  geom_cladelab(node = 257, label = "GSH01",
                # textcolor = "purple3",
                barcolor = "purple3",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.005, offset.text = 0.0002) +
  geom_cladelab(node = 242, label = "GSH09b",
                # textcolor = "gold2",
                barcolor = "gold2",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.007, offset.text = 0.0002) +
  geom_cladelab(node = 218, label = "GSH09c_ETP",
                # textcolor = "darkorange",
                barcolor = "darkorange",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.0085, offset.text = 0.0002) +
  geom_cladelab(node = 149, label = "GSH13c_ETP",
                # textcolor = "turquoise2",
                barcolor = "turquoise2",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.008, offset.text = 0.0002) +
  
  # label clades - TEP nominal spp.
  geom_cladelab(node = 257, label = "P. effusa", fontsize = 3, fontface = 3, offset = 0.01, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 242, label = "P. meandrina", fontsize = 3, fontface = 3, offset = 0.013, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 218, label = "P. capitata", fontsize = 3, fontface = 3, offset = 0.0155, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 149, label = "P. lacera", fontsize = 3, fontface = 3, offset = 0.016, offset.text = 0.0004, barcolor = NA) +
  # label clades - absent nominal spp.
  geom_cladelab(node = 197, label = "P. damicornis", fontsize = 3, fontface = 3, offset = 0.012, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 171, label = "P. acuta", fontsize = 3, fontface = 3, offset = 0.016, offset.text = 0.0004, barcolor = NA) +
  
  

  # setup tree visualization
  geom_rootedge(rootedge = 0.001) +
  geom_treescale(width = 0.01, x = 0, y = -1, linesize = 0.75) +
  ggplot2::xlim(-0.001, 0.05) +
  scale_color_manual(values = c("red", "navyblue", "black")) +
  scale_size_manual(values = c(1.85, 1.85, 1.5, 1.5)) +
  theme(plot.margin = margin(2,2,2,2, unit = "mm"),
        legend.position = "none",
        plot.title = element_blank(),
        plot.subtitle = element_text(size = 10))

#
pdf("./outputs/figures/Fig2_iqtree_rep_50pct_tree_annotated_revised.pdf", width = 6.5, height = 7)
print(p_annotated)
dev.off()
```

### Figure S11
#### 75% matrix
 138 taxa, 1,263 UCE loci, 613,088 bp, 166,742  informative sites
```{r import_tree}
tree <- read.newick("./outputs/iqtree/iqtree_representative_5X_phyluce_iqtree_75.treefile")

# midpoint root
tree <- midpoint.root(tree)

# drop outgroup tips (easier to view node labels)
tree <- ape::drop.tip(tree, tip = "Stylophora_pistillata_ANT17_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052307_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052309_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052305_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052306_contigs")

# set root-edge
tree$root.edge <- 1
# check topology
ggtree(tree)
```
```{r adjust_tip_labels}
# adjust tip labels
# remove contigs suffix 
tips <- gsub("_contigs$", "", tree$tip.label)
# remove sample number suffix 
tips <- gsub("_S[[:digit:]]*$", "", tips)

# Reference genomes: tip names
tips <- gsub("Stylophora_", "S. ", tips)
tips <- gsub("_ANT17", " ANT17", tips)
tips <- gsub("pocillopora_", "P. ", tips)
tips <- gsub("_G", " G", tips)
tips <- gsub("_SAMN", " SAMN", tips)

# Oury samples:
oury_tip_order <- tips[tips %in% oury_samples$NCBI_SSR]

tips[tips %in% oury_samples$NCBI_SSR] <- oury_samples[match(oury_tip_order, oury_samples$NCBI_SSR),]$tip_label

# need to change reference genome labels to include GenBank accessions
# Stephens et al. 2022 genomes do not have GenBank/RefSeq accessions, using BioSample numbers
# SAMN21845732 - P. meandrina
# SAMN22898959 - P. acuta
```
```{r check_tree_tips}
# check tips
tips
# assign new tip labels to tree
tree$tip.label <- tips

# basic viz
gg_tree(tree)
```
```{r tip_metadata}
# create dataframe holding metadata 
tips_metadata <- data.frame(tips)
tips_metadata$class <- ifelse(grepl("Glo", tips), "Outgroup", ifelse(grepl("P\\.|S\\.", tips), "Genome", ifelse(grepl("GSH", tips), "Oury", "Connelly")))
# red for USNM, dark blue for genomes, dark green for Oury, black for Connelly
```
```{r node_labels}
# node labels with bootstrap support
node.labels <- tree$node.label
# preserve original labels
node.labels.original <- node.labels
node.labels <- node.labels.original

# remove root label
node.labels[node.labels == "Root"] <- ""
# remove 100% supported node labels
# node.labels[node.labels == "100/100"] <- ""
# define high-support bootstrap values - >90%?
node.labels[grepl("^100", node.labels)] <- ""
node.labels[grepl("^9", node.labels)] <- ""
# mark low-support bootstrap values with asterisks - <80%?
# node.labels[grepl("^[8]", node.labels)] <- ""
# node.labels[grepl("^[5-7]", node.labels)] <- "<80"
# node.labels[grepl("^[0-4]", node.labels)] <- "<50"

# check to see if formatting is correct
node.labels
#
tree$node.label <- node.labels
```
```{r full_ggtree_viz}
# pdf("./outputs/figures/Fig2_iqtree_rep_50pct_tree.pdf", width = 6.5, height = 8.5)
#
tree_title <- expression(paste("IQ-TREE Maximum Likelihood Tree with reference ", italic("Pocillopora"), " GSH samples"))
loci_info <- c()# c("50% sample occupancy, 1,519 UCE loci, 713,305 bp, 193,855 informative sites")
#

p <- tree %>% ggtree(., right = TRUE) %<+% tips_metadata + 
  
  # highlight clades
  geom_highlight(node = 218, fill = "darkorange", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 147, fill = "turquoise2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 257, fill = "purple3", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 242, fill = "gold2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +

  # node labels (numeric)
  geom_text(aes(label=node), hjust=-.3, size = 1) +
  
  # node labels
  geom_nodelab(size = 2, hjust = 0, nudge_y = 0, geom = "shadowtext", color = "grey90", bg.color = "white") +
  
  # tip labels
  geom_tiplab(hjust = 0, aes(color = `class`, size = `class`)) +
  
  # label clades
  geom_cladelab(node = 257, label = "P. effusa", fontsize = 3, fontface = 3, offset = 0.008, offset.text = 0.0004) +
  # 
  geom_cladelab(node = 242, label = "P. meandrina", fontsize = 3, fontface = 3, offset = 0.009, offset.text = 0.0004) +
  # 
  geom_cladelab(node = 218, label = "P. capitata", fontsize = 3, fontface = 3, offset = 0.011, offset.text = 0.0004) +
  # 
  geom_cladelab(node = 147, label = "P. lacera", fontsize = 3, fontface = 3, offset = 0.011, offset.text = 0.0004) +

  # setup tree visualization
  geom_rootedge(rootedge = 0.001) +
  geom_treescale(width = 0.01, x = 0, y = -1, linesize = 0.75) +
  ggplot2::xlim(-0.001, 0.06) +
  scale_color_manual(values = c("red", "navyblue", "black")) +
  scale_size_manual(values = c(1.75, 1.75, 1.5, 1.5)) +
  theme(plot.margin = margin(5,2,2,5, unit = "mm"),
        legend.position = "none",
        plot.title = element_blank(),
        plot.subtitle = element_text(size = 10)) +
  ggtitle(tree_title, subtitle = loci_info)
#

print(p)
# dev.off()
```
```{r custom_collapsed_annotated_tree}
p <- tree %>% ggtree(., right = TRUE) %<+% tips_metadata 

p_collapsed <- p %>%
  
  # # scale clades down
  scaleClade(., node = 171, scale = 0.25) %>%
  # scaleClade(., node = 194, scale = 1) %>%
  # 
  # # scale clades up
  scaleClade(., node = 257, scale = 1.4) %>%
  scaleClade(., node = 242, scale = 1.4) %>%
  scaleClade(., node = 216, scale = 1.4) %>%
  scaleClade(., node = 144, scale = 1.4) %>%
   # scaleClade(., node = 146, scale = 1.4) %>%


  # collapse clades
  collapse(140, 'min', fill='white', color="black", alpha=.4) %>% # GSH15
  collapse(158, 'min', fill='white', color="black", alpha=.4) %>% # GSH13a
  collapse(162, 'min', fill='white', color="black", alpha=.4) %>% # GSH12
  collapse(166, 'min', fill='white', color="black", alpha=.4) %>% # GSH14
  collapse(171, 'min', fill='white', color="black", alpha=.4) %>% # GSH05
  # 
  # # collapse(194, 'min', fill='white', color="black", alpha=.4) %>%
  collapse(200, 'min', fill='white', color="black", alpha=.4) %>% # GSH04c
  collapse(201, 'min', fill='white', color="black", alpha=.4) %>% # GSH10
  collapse(206, 'min', fill='white', color="black", alpha=.4) %>% # GSH04b
  collapse(210, 'min', fill='white', color="black", alpha=.4) %>% # GSH04a
  # 
  collapse(237, 'min', fill='white', color="black", alpha=.4) %>% # GSH09a
  collapse(253, 'min', fill='white', color="black", alpha=.4) # GSH13b


p_annotated <- p_collapsed +
  
  # highlight clades
  geom_highlight(node = 218, fill = "darkorange", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 147, fill = "turquoise2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 259, fill = "purple3", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 245, fill = "gold2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 56, fill = "gold2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +

  # node labels (numeric)
  # geom_text(aes(label=node), hjust=-.3, size = 1) +
  
  # node labels
  geom_nodelab(size = 2, hjust = 0, nudge_y = 0, geom = "shadowtext", color = "grey90", bg.color = "white") +
  
  # tip labels
  geom_tiplab(hjust = 0, aes(color = `class`, size = `class`)) +
  
  # label clades - collapsed GSHs
  # clade 2
  geom_cladelab(node = 237, label = "GSH09a", fontsize = 2, fontface = 1, offset = 0.008, offset.text = 0.0001) +
  geom_cladelab(node = 253, label = "GSH13b", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  # # clade 3
  geom_cladelab(node = 140, label = "GSH15", fontsize = 2, fontface = 1, offset = 0.0065, offset.text = 0.0001) +
  geom_cladelab(node = 158, label = "GSH13a", fontsize = 2, fontface = 1, offset = 0.007, offset.text = 0.0001) +
  geom_cladelab(node = 162, label = "GSH12", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  geom_cladelab(node = 166, label = "GSH14", fontsize = 2, fontface = 1, offset = 0.009, offset.text = 0.0001) +
  # # clade 4
  geom_cladelab(node = 200, label = "GSH04c", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  geom_cladelab(node = 201, label = "GSH10", fontsize = 2, fontface = 1, offset = 0.005, offset.text = 0.0001) +
  geom_cladelab(node = 206, label = "GSH04b", fontsize = 2, fontface = 1, offset = 0.006, offset.text = 0.0001) +
  geom_cladelab(node = 210, label = "GSH04a", fontsize = 2, fontface = 1, offset = 0.007, offset.text = 0.0001) +
  # # 
  geom_cladelab(node = 171, label = "GSH05", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  # #
  
  # label clades - TEP GSHs
  geom_cladelab(node = 257, label = "GSH01",
                # textcolor = "purple3",
                barcolor = "purple3",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.005, offset.text = 0.0002) +
  geom_cladelab(node = 242, label = "GSH09b",
                # textcolor = "gold2",
                barcolor = "gold2",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.007, offset.text = 0.0002) +
  geom_cladelab(node = 218, label = "GSH09c_ETP",
                # textcolor = "darkorange",
                barcolor = "darkorange",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.0085, offset.text = 0.0002) +
  geom_cladelab(node = 147, label = "GSH13c_ETP",
                # textcolor = "turquoise2",
                barcolor = "turquoise2",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.008, offset.text = 0.0002) +
  
  # label clades - TEP nominal spp.
  geom_cladelab(node = 257, label = "P. effusa", fontsize = 3, fontface = 3, offset = 0.01, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 242, label = "P. meandrina", fontsize = 3, fontface = 3, offset = 0.013, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 218, label = "P. capitata", fontsize = 3, fontface = 3, offset = 0.0155, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 147, label = "P. lacera", fontsize = 3, fontface = 3, offset = 0.016, offset.text = 0.0004, barcolor = NA) +
  # label clades - absent nominal spp.
  geom_cladelab(node = 197, label = "P. damicornis", fontsize = 3, fontface = 3, offset = 0.012, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 171, label = "P. acuta", fontsize = 3, fontface = 3, offset = 0.016, offset.text = 0.0004, barcolor = NA) +
  
  

  # setup tree visualization
  geom_rootedge(rootedge = 0.001) +
  geom_treescale(width = 0.01, x = 0, y = -1, linesize = 0.75) +
  ggplot2::xlim(-0.001, 0.05) +
  scale_color_manual(values = c("red", "navyblue", "black")) +
  scale_size_manual(values = c(1.85, 1.85, 1.5, 1.5)) +
  theme(plot.margin = margin(2,2,2,2, unit = "mm"),
        legend.position = "none",
        plot.title = element_blank(),
        plot.subtitle = element_text(size = 10))

#
pdf("./outputs/figures/FigS11_iqtree_rep_75pct_tree_annotated_revised.pdf", width = 6.5, height = 7)
print(p_annotated)
dev.off()
```


### PociGSkim NGSAdmix K=5 top 5 reps (29) + Oury GSH top 5 reps (103) + reference genome samples (8) + outgroups (1) + Isla del Coco mtORF 3b samples (5) (n = 143 samples)
#### 75% matrix
 143 taxa, 1,273 UCE loci, 613,550 bp, 167,666  informative sites
```{r import_tree}
tree <- read.newick("./outputs/iqtree/iqtree_representative_5X_plus3b_phyluce_iqtree_75.treefile")

# midpoint root
tree <- midpoint.root(tree)

# drop outgroup tips (easier to view node labels)
tree <- ape::drop.tip(tree, tip = "Stylophora_pistillata_ANT17_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052307_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052309_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052305_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052306_contigs")

# set root-edge
tree$root.edge <- 1
# check topology
ggtree(tree)
```
```{r adjust_tip_labels}
# adjust tip labels
# remove contigs suffix 
tips <- gsub("_contigs$", "", tree$tip.label)
# remove sample number suffix 
tips <- gsub("_S[[:digit:]]*$", "", tips)

# Reference genomes: tip names
tips <- gsub("Stylophora_", "S. ", tips)
tips <- gsub("_ANT17", " ANT17", tips)
tips <- gsub("pocillopora_", "P. ", tips)
tips <- gsub("_G", " G", tips)
tips <- gsub("_SAMN", " SAMN", tips)

# Oury samples:
oury_tip_order <- tips[tips %in% oury_samples$NCBI_SSR]

tips[tips %in% oury_samples$NCBI_SSR] <- oury_samples[match(oury_tip_order, oury_samples$NCBI_SSR),]$tip_label

# need to change reference genome labels to include GenBank accessions
# Stephens et al. 2022 genomes do not have GenBank/RefSeq accessions, using BioSample numbers
# SAMN21845732 - P. meandrina
# SAMN22898959 - P. acuta
```
```{r check_tree_tips}
# check tips
tips
# assign new tip labels to tree
tree$tip.label <- tips

# basic viz
gg_tree(tree)
```
```{r tip_metadata}
# create dataframe holding metadata 
tips_metadata <- data.frame(tips)
tips_metadata$class <- ifelse(grepl("Glo", tips), "Outgroup", ifelse(grepl("P\\.|S\\.", tips), "Genome", ifelse(grepl("GSH", tips), "Oury", "Connelly")))
# red for USNM, dark blue for genomes, dark green for Oury, black for Connelly
```
```{r node_labels}
# node labels with bootstrap support
node.labels <- tree$node.label
# preserve original labels
node.labels.original <- node.labels
node.labels <- node.labels.original

# remove root label
node.labels[node.labels == "Root"] <- ""
# remove 100% supported node labels
# node.labels[node.labels == "100/100"] <- ""
# define high-support bootstrap values - >90%?
node.labels[grepl("^100", node.labels)] <- ""
node.labels[grepl("^9", node.labels)] <- ""
# mark low-support bootstrap values with asterisks - <80%?
# node.labels[grepl("^[8]", node.labels)] <- ""
# node.labels[grepl("^[5-7]", node.labels)] <- "<80"
# node.labels[grepl("^[0-4]", node.labels)] <- "<50"

# check to see if formatting is correct
node.labels
#
tree$node.label <- node.labels
```
```{r full_ggtree_viz}
pdf("./outputs/figures/iqtree_75pct_tree.pdf", width = 6.5, height = 8.5)
#
tree_title <- expression(paste("IQ-TREE Maximum Likelihood Tree with reference ", italic("Pocillopora"), " GSH samples"))
loci_info <- c("75% sample occupancy, 1,273 UCE loci, 613,550 bp, 167,666 informative sites")
#

tree %>% ggtree(., right = TRUE) %<+% tips_metadata + 
  # figure out way to highlight clades
  # geom_highlight(node = 69, fill = "orange", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  # geom_highlight(node = 115, fill = "turquoise", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  # geom_highlight(node = 105, fill = "magenta", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  # geom_highlight(node = 62, fill = "purple", type = "gradient", gradient.direction = 'rt', alpha = .6) +
  # geom_highlight(node = 70, fill = "purple", type = "gradient", gradient.direction = 'rt', alpha = .6) +

  # node labels (numeric)
  # geom_text(aes(label=node), hjust=-.3, size = 1) +
  
  # node labels
  geom_nodelab(size = 2, hjust = 0, nudge_y = 0, geom = "shadowtext", color = "grey90", bg.color = "white") +
  
  # tip labels
  geom_tiplab(hjust = 0, aes(color = `class`, size = `class`)) +
  
  # label clades
  # geom_cladelab(node = 41, label = "P. effusa", fontsize = 3, fontface = 3, offset = 0.01, offset.text = 0.0004) +
  # geom_cladelab(node = 47, label = "P. grandis", fontsize = 3, fontface = 3, offset = 0.011, offset.text = 0.0004) +
  # geom_cladelab(node = 60, label = "P. meandrina", fontsize = 3, fontface = 3, offset = 0.014, offset.text = 0.0004) +
  # geom_cladelab(node = 62, label = "P. acuta / damicornis", fontsize = 3, fontface = 3, offset = 0.009, offset.text = 0.0004) +
  # geom_cladelab(node = 70, label = "P. verrucosa", fontsize = 3, fontface = 3, offset = 0.011, offset.text = 0.0004) +

  # setup tree visualization
  geom_rootedge(rootedge = 0.001) +
  geom_treescale(width = 0.01, x = 0, y = -1, linesize = 0.75) +
  ggplot2::xlim(-0.001, 0.06) +
  scale_color_manual(values = c("red", "navyblue", "black")) +
  scale_size_manual(values = c(1.75, 1.75, 1.5, 1.5)) +
  theme(plot.margin = margin(5,2,2,5, unit = "mm"),
        legend.position = "none",
        plot.title = element_blank(),
        plot.subtitle = element_text(size = 10)) +
  ggtitle(tree_title, subtitle = loci_info)
#
dev.off()
```

#### 50% matrix
 143 taxa
```{r import_tree}
tree <- read.newick("./outputs/iqtree/iqtree_representative_5X_plus3b_phyluce_iqtree_50.treefile")

# midpoint root
tree <- midpoint.root(tree)

# drop outgroup tips (easier to view node labels)
tree <- ape::drop.tip(tree, tip = "Stylophora_pistillata_ANT17_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052307_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052309_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052305_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052306_contigs")

# set root-edge
tree$root.edge <- 1
# check topology
ggtree(tree)
```
```{r adjust_tip_labels}
# adjust tip labels
# remove contigs suffix 
tips <- gsub("_contigs$", "", tree$tip.label)
# remove sample number suffix 
tips <- gsub("_S[[:digit:]]*$", "", tips)

# Reference genomes: tip names
tips <- gsub("Stylophora_", "S. ", tips)
tips <- gsub("_ANT17", " ANT17", tips)
tips <- gsub("pocillopora_", "P. ", tips)
tips <- gsub("_G", " G", tips)
tips <- gsub("_SAMN", " SAMN", tips)

# Oury samples:
oury_tip_order <- tips[tips %in% oury_samples$NCBI_SSR]

tips[tips %in% oury_samples$NCBI_SSR] <- oury_samples[match(oury_tip_order, oury_samples$NCBI_SSR),]$tip_label

# need to change reference genome labels to include GenBank accessions
# Stephens et al. 2022 genomes do not have GenBank/RefSeq accessions, using BioSample numbers
# SAMN21845732 - P. meandrina
# SAMN22898959 - P. acuta
```
```{r check_tree_tips}
# check tips
tips
# assign new tip labels to tree
tree$tip.label <- tips

# basic viz
gg_tree(tree)
```
```{r tip_metadata}
# create dataframe holding metadata 
tips_metadata <- data.frame(tips)
tips_metadata$class <- ifelse(grepl("Glo", tips), "Outgroup", ifelse(grepl("P\\.|S\\.", tips), "Genome", ifelse(grepl("GSH", tips), "Oury", "Connelly")))
# red for USNM, dark blue for genomes, dark green for Oury, black for Connelly
```
```{r node_labels}
# node labels with bootstrap support
node.labels <- tree$node.label
# preserve original labels
node.labels.original <- node.labels
node.labels <- node.labels.original

# remove root label
node.labels[node.labels == "Root"] <- ""
# remove 100% supported node labels
# node.labels[node.labels == "100/100"] <- ""
# define high-support bootstrap values - >90%?
node.labels[grepl("^100", node.labels)] <- ""
node.labels[grepl("^9", node.labels)] <- ""
# mark low-support bootstrap values with asterisks - <80%?
# node.labels[grepl("^[8]", node.labels)] <- ""
# node.labels[grepl("^[5-7]", node.labels)] <- "<80"
# node.labels[grepl("^[0-4]", node.labels)] <- "<50"

# check to see if formatting is correct
node.labels
#
tree$node.label <- node.labels
```
```{r full_ggtree_viz}
pdf("./outputs/figures/iqtree_plus3b_50pct_tree.pdf", width = 6.5, height = 8.5)
#
tree_title <- expression(paste("IQ-TREE Maximum Likelihood Tree with reference ", italic("Pocillopora"), " GSH samples"))
loci_info <- "" #c("50% sample occupancy, 1,519 UCE loci, 713,305 bp, 193,855 informative sites")
#

tree %>% ggtree(., right = TRUE) %<+% tips_metadata + 
  # figure out way to highlight clades
  # geom_highlight(node = 69, fill = "orange", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  # geom_highlight(node = 115, fill = "turquoise", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  # geom_highlight(node = 105, fill = "magenta", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  # geom_highlight(node = 62, fill = "purple", type = "gradient", gradient.direction = 'rt', alpha = .6) +
  # geom_highlight(node = 70, fill = "purple", type = "gradient", gradient.direction = 'rt', alpha = .6) +

  # node labels (numeric)
  geom_text(aes(label=node), hjust=-.3, size = 1) +
  
  # node labels
  geom_nodelab(size = 2, hjust = 0, nudge_y = 0, geom = "shadowtext", color = "grey90", bg.color = "white") +
  
  # tip labels
  geom_tiplab(hjust = 0, aes(color = `class`, size = `class`)) +
  
  # label clades
  # geom_cladelab(node = 41, label = "P. effusa", fontsize = 3, fontface = 3, offset = 0.01, offset.text = 0.0004) +
  # geom_cladelab(node = 47, label = "P. grandis", fontsize = 3, fontface = 3, offset = 0.011, offset.text = 0.0004) +
  # geom_cladelab(node = 60, label = "P. meandrina", fontsize = 3, fontface = 3, offset = 0.014, offset.text = 0.0004) +
  # geom_cladelab(node = 62, label = "P. acuta / damicornis", fontsize = 3, fontface = 3, offset = 0.009, offset.text = 0.0004) +
  # geom_cladelab(node = 70, label = "P. verrucosa", fontsize = 3, fontface = 3, offset = 0.011, offset.text = 0.0004) +

  # setup tree visualization
  geom_rootedge(rootedge = 0.001) +
  geom_treescale(width = 0.01, x = 0, y = -1, linesize = 0.75) +
  ggplot2::xlim(-0.001, 0.06) +
  scale_color_manual(values = c("red", "navyblue", "black")) +
  scale_size_manual(values = c(1.75, 1.75, 1.5, 1.5)) +
  theme(plot.margin = margin(5,2,2,5, unit = "mm"),
        legend.position = "none",
        plot.title = element_blank(),
        plot.subtitle = element_text(size = 10)) +
  ggtitle(tree_title, subtitle = loci_info)
#
dev.off()
```
### Figure S12
#### 50% matrix
 143 taxa, 1,521 UCE loci, 710,682 bp, 192,776  informative sites
```{r custom_collapsed_annotated_tree}
p <- tree %>% ggtree(., right = TRUE) %<+% tips_metadata 

p_collapsed <- p %>%
  
  # # scale clades down
  scaleClade(., node = 181, scale = 0.25) %>%
  # scaleClade(., node = 194, scale = 1) %>%
  # 
  # # scale clades up
  scaleClade(., node = 267, scale = 1.4) %>%
  scaleClade(., node = 252, scale = 1.4) %>%
  scaleClade(., node = 226, scale = 1.4) %>%
  scaleClade(., node = 143, scale = 1.4) %>%
   # scaleClade(., node = 152, scale = 1.4) %>%


  # collapse clades
  collapse(144, 'min', fill='white', color="black", alpha=.4) %>% # GSH15
  collapse(164, 'min', fill='white', color="black", alpha=.4) %>% # GSH13a
  collapse(168, 'min', fill='white', color="black", alpha=.4) %>% # GSH12
  collapse(176, 'min', fill='white', color="black", alpha=.4) %>% # GSH14
  collapse(181, 'min', fill='white', color="black", alpha=.4) %>% # GSH05
  # # 
  # # collapse(194, 'min', fill='white', color="black", alpha=.4) %>%
  collapse(210, 'min', fill='white', color="black", alpha=.4) %>% # GSH04c
  collapse(211, 'min', fill='white', color="black", alpha=.4) %>% # GSH10
  collapse(216, 'min', fill='white', color="black", alpha=.4) %>% # GSH04b
  collapse(220, 'min', fill='white', color="black", alpha=.4) %>% # GSH04a
  # # 
  collapse(247, 'min', fill='white', color="black", alpha=.4) %>% # GSH09a
  collapse(263, 'min', fill='white', color="black", alpha=.4) # GSH13b


p_annotated <- p_collapsed +
  
  # highlight clades
  geom_highlight(node = 228, fill = "darkorange", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 155, fill = "turquoise2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 269, fill = "purple3", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 255, fill = "gold2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 72, fill = "gold2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +

  # node labels (numeric)
  # geom_text(aes(label=node), hjust=-.3, size = 1) +
  
  # node labels
  geom_nodelab(size = 2, hjust = 0, nudge_y = 0, geom = "shadowtext", color = "grey90", bg.color = "white") +
  
  # tip labels
  geom_tiplab(hjust = 0, aes(color = `class`, size = `class`)) +
  
  # label clades - collapsed GSHs
  # # clade 2
  geom_cladelab(node = 247, label = "GSH09a", fontsize = 2, fontface = 1, offset = 0.008, offset.text = 0.0001) +
  geom_cladelab(node = 263, label = "GSH13b", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  # # clade 3
  geom_cladelab(node = 144, label = "GSH15", fontsize = 2, fontface = 1, offset = 0.0065, offset.text = 0.0001) +
  geom_cladelab(node = 164, label = "GSH13a", fontsize = 2, fontface = 1, offset = 0.007, offset.text = 0.0001) +
  geom_cladelab(node = 168, label = "GSH12", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  geom_cladelab(node = 176, label = "GSH14", fontsize = 2, fontface = 1, offset = 0.009, offset.text = 0.0001) +
  # # clade 4
  geom_cladelab(node = 210, label = "GSH04c", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  geom_cladelab(node = 211, label = "GSH10", fontsize = 2, fontface = 1, offset = 0.005, offset.text = 0.0001) +
  geom_cladelab(node = 216, label = "GSH04b", fontsize = 2, fontface = 1, offset = 0.006, offset.text = 0.0001) +
  geom_cladelab(node = 220, label = "GSH04a", fontsize = 2, fontface = 1, offset = 0.007, offset.text = 0.0001) +
  # #
  geom_cladelab(node = 181, label = "GSH05", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  # # #
  # 
  # label clades - TEP GSHs
  geom_cladelab(node = 267, label = "GSH01",
                # textcolor = "purple3",
                barcolor = "purple3",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.005, offset.text = 0.0002) +
  geom_cladelab(node = 252, label = "GSH09b",
                # textcolor = "gold2",
                barcolor = "gold2",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.007, offset.text = 0.0002) +
  geom_cladelab(node = 228, label = "GSH09c_ETP",
                # textcolor = "darkorange",
                barcolor = "darkorange",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.0085, offset.text = 0.0002) +
  geom_cladelab(node = 155, label = "GSH13c_ETP",
                # textcolor = "turquoise2",
                barcolor = "turquoise2",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.008, offset.text = 0.0002) +
  
  # label clades - TEP nominal spp.
  geom_cladelab(node = 267, label = "P. effusa", fontsize = 3, fontface = 3, offset = 0.01, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 252, label = "P. meandrina", fontsize = 3, fontface = 3, offset = 0.013, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 228, label = "P. capitata", fontsize = 3, fontface = 3, offset = 0.0155, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 155, label = "P. lacera", fontsize = 3, fontface = 3, offset = 0.016, offset.text = 0.0004, barcolor = NA) +
  # label clades - absent nominal spp.
  geom_cladelab(node = 207, label = "P. damicornis", fontsize = 3, fontface = 3, offset = 0.012, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 181, label = "P. acuta", fontsize = 3, fontface = 3, offset = 0.016, offset.text = 0.0004, barcolor = NA) +
  # 
  

  # setup tree visualization
  geom_rootedge(rootedge = 0.001) +
  geom_treescale(width = 0.01, x = 0, y = -12, linesize = 0.75) +
  ggplot2::xlim(-0.001, 0.05) +
  scale_color_manual(values = c("red", "navyblue", "black")) +
  scale_size_manual(values = c(1.85, 1.85, 1.5, 1.5)) +
  theme(plot.margin = margin(2,2,2,2, unit = "mm"),
        legend.position = "none",
        plot.title = element_blank(),
        plot.subtitle = element_text(size = 10))

#
print(p_annotated)
```
```{r}
pdf("./outputs/figures/FigS12_iqtree_rep_plus3b_50pct_tree_annotated.pdf", width = 6.5, height = 7)
print(p_annotated)
dev.off()
```

### Figure S13
#### 75% matrix
143 taxa, 1,252 UCE loci, 599,496 bp, 162,250  informative sites
```{r import_tree}
tree <- read.newick("./outputs/iqtree/iqtree_representative_5X_plus3b_phyluce_iqtree_75.treefile")

# midpoint root
tree <- midpoint.root(tree)

# drop outgroup tips (easier to view node labels)
tree <- ape::drop.tip(tree, tip = "Stylophora_pistillata_ANT17_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052307_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052309_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052305_contigs")
tree <- ape::drop.tip(tree, tip = "SRR19052306_contigs")

# set root-edge
tree$root.edge <- 1
# check topology
ggtree(tree)
```
```{r adjust_tip_labels}
# adjust tip labels
# remove contigs suffix 
tips <- gsub("_contigs$", "", tree$tip.label)
# remove sample number suffix 
tips <- gsub("_S[[:digit:]]*$", "", tips)

# Reference genomes: tip names
tips <- gsub("Stylophora_", "S. ", tips)
tips <- gsub("_ANT17", " ANT17", tips)
tips <- gsub("pocillopora_", "P. ", tips)
tips <- gsub("_G", " G", tips)
tips <- gsub("_SAMN", " SAMN", tips)

# Oury samples:
oury_tip_order <- tips[tips %in% oury_samples$NCBI_SSR]

tips[tips %in% oury_samples$NCBI_SSR] <- oury_samples[match(oury_tip_order, oury_samples$NCBI_SSR),]$tip_label

# need to change reference genome labels to include GenBank accessions
# Stephens et al. 2022 genomes do not have GenBank/RefSeq accessions, using BioSample numbers
# SAMN21845732 - P. meandrina
# SAMN22898959 - P. acuta
```
```{r check_tree_tips}
# check tips
tips
# assign new tip labels to tree
tree$tip.label <- tips

# basic viz
gg_tree(tree)
```
```{r tip_metadata}
# create dataframe holding metadata 
tips_metadata <- data.frame(tips)
tips_metadata$class <- ifelse(grepl("Glo", tips), "Outgroup", ifelse(grepl("P\\.|S\\.", tips), "Genome", ifelse(grepl("GSH", tips), "Oury", "Connelly")))
# red for USNM, dark blue for genomes, dark green for Oury, black for Connelly
```
```{r node_labels}
# node labels with bootstrap support
node.labels <- tree$node.label
# preserve original labels
node.labels.original <- node.labels
node.labels <- node.labels.original

# remove root label
node.labels[node.labels == "Root"] <- ""
# remove 100% supported node labels
# node.labels[node.labels == "100/100"] <- ""
# define high-support bootstrap values - >90%?
node.labels[grepl("^100", node.labels)] <- ""
node.labels[grepl("^9", node.labels)] <- ""
# mark low-support bootstrap values with asterisks - <80%?
# node.labels[grepl("^[8]", node.labels)] <- ""
# node.labels[grepl("^[5-7]", node.labels)] <- "<80"
# node.labels[grepl("^[0-4]", node.labels)] <- "<50"

# check to see if formatting is correct
node.labels
#
tree$node.label <- node.labels
```
```{r full_ggtree_viz}
# pdf("./outputs/figures/Fig2_iqtree_rep_50pct_tree.pdf", width = 6.5, height = 8.5)
#
tree_title <- expression(paste("IQ-TREE Maximum Likelihood Tree with reference ", italic("Pocillopora"), " GSH samples"))
loci_info <- c()# c("50% sample occupancy, 1,519 UCE loci, 713,305 bp, 193,855 informative sites")
#

p <- tree %>% ggtree(., right = TRUE) %<+% tips_metadata + 
  
  # highlight clades
  geom_highlight(node = 227, fill = "darkorange", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 158, fill = "turquoise2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 267, fill = "purple3", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 248, fill = "gold2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +

  # node labels (numeric)
  geom_text(aes(label=node), hjust=-.3, size = 1) +
  
  # node labels
  geom_nodelab(size = 2, hjust = 0, nudge_y = 0, geom = "shadowtext", color = "grey90", bg.color = "white") +
  
  # tip labels
  geom_tiplab(hjust = 0, aes(color = `class`, size = `class`)) +
  
  # label clades
  geom_cladelab(node = 267, label = "P. effusa", fontsize = 3, fontface = 3, offset = 0.008, offset.text = 0.0004) +
  # 
  geom_cladelab(node = 248, label = "P. meandrina", fontsize = 3, fontface = 3, offset = 0.009, offset.text = 0.0004) +
  # 
  geom_cladelab(node = 227, label = "P. capitata", fontsize = 3, fontface = 3, offset = 0.011, offset.text = 0.0004) +
  # 
  geom_cladelab(node = 158, label = "P. lacera", fontsize = 3, fontface = 3, offset = 0.011, offset.text = 0.0004) +

  # setup tree visualization
  geom_rootedge(rootedge = 0.001) +
  geom_treescale(width = 0.01, x = 0, y = -1, linesize = 0.75) +
  ggplot2::xlim(-0.001, 0.06) +
  scale_color_manual(values = c("red", "navyblue", "black")) +
  scale_size_manual(values = c(1.75, 1.75, 1.5, 1.5)) +
  theme(plot.margin = margin(5,2,2,5, unit = "mm"),
        legend.position = "none",
        plot.title = element_blank(),
        plot.subtitle = element_text(size = 10)) +
  ggtitle(tree_title, subtitle = loci_info)
#

print(p)
# dev.off()
```
```{r custom_collapsed_annotated_tree}
p <- tree %>% ggtree(., right = TRUE) %<+% tips_metadata 

p_collapsed <- p %>%
  
  # # scale clades down
  scaleClade(., node = 181, scale = 0.25) %>%
  # scaleClade(., node = 194, scale = 1) %>%
  # 
  # # scale clades up
  scaleClade(., node = 267, scale = 1.4) %>%
  scaleClade(., node = 248, scale = 1.4) %>%
  scaleClade(., node = 225, scale = 1.4) %>%
  scaleClade(., node = 153, scale = 1.4) %>%
  scaleClade(., node = 174, scale = 1.4) %>%


  # collapse clades
  collapse(147, 'min', fill='white', color="black", alpha=.4) %>% # GSH15
  collapse(166, 'min', fill='white', color="black", alpha=.4) %>% # GSH13a
  collapse(170, 'min', fill='white', color="black", alpha=.4) %>% # GSH12
  collapse(176, 'min', fill='white', color="black", alpha=.4) %>% # GSH14
  collapse(181, 'min', fill='white', color="black", alpha=.4) %>% # GSH05
  # 
  # # collapse(194, 'min', fill='white', color="black", alpha=.4) %>%
  collapse(210, 'min', fill='white', color="black", alpha=.4) %>% # GSH04c
  collapse(211, 'min', fill='white', color="black", alpha=.4) %>% # GSH10
  collapse(216, 'min', fill='white', color="black", alpha=.4) %>% # GSH04b
  collapse(220, 'min', fill='white', color="black", alpha=.4) %>% # GSH04a
  # 
  collapse(263, 'min', fill='white', color="black", alpha=.4) %>% # GSH09a
  collapse(259, 'min', fill='white', color="black", alpha=.4) # GSH13b


p_annotated <- p_collapsed +
  
  # highlight clades
  geom_highlight(node = 227, fill = "darkorange", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 158, fill = "turquoise2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 269, fill = "purple3", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 251, fill = "gold2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +
  geom_highlight(node = 61, fill = "gold2", type = "gradient", gradient.direction = 'rt', alpha = 1, to.bottom = T) +

  # node labels (numeric)
  # geom_text(aes(label=node), hjust=-.3, size = 1) +
  
  # node labels
  geom_nodelab(size = 2, hjust = 0, nudge_y = 0, geom = "shadowtext", color = "grey90", bg.color = "white") +
  
  # tip labels
  geom_tiplab(hjust = 0, aes(color = `class`, size = `class`)) +
  
  # label clades - collapsed GSHs
  # clade 2
  geom_cladelab(node = 263, label = "GSH09a", fontsize = 2, fontface = 1, offset = 0.008, offset.text = 0.0001) +
  geom_cladelab(node = 259, label = "GSH13b", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  # # clade 3
  geom_cladelab(node = 147, label = "GSH15", fontsize = 2, fontface = 1, offset = 0.0065, offset.text = 0.0001) +
  geom_cladelab(node = 166, label = "GSH13a", fontsize = 2, fontface = 1, offset = 0.007, offset.text = 0.0001) +
  geom_cladelab(node = 170, label = "GSH12", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  geom_cladelab(node = 176, label = "GSH14", fontsize = 2, fontface = 1, offset = 0.009, offset.text = 0.0001) +
  # # clade 4
  geom_cladelab(node = 210, label = "GSH04c", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  geom_cladelab(node = 211, label = "GSH10", fontsize = 2, fontface = 1, offset = 0.005, offset.text = 0.0001) +
  geom_cladelab(node = 216, label = "GSH04b", fontsize = 2, fontface = 1, offset = 0.006, offset.text = 0.0001) +
  geom_cladelab(node = 220, label = "GSH04a", fontsize = 2, fontface = 1, offset = 0.007, offset.text = 0.0001) +
  # # 
  geom_cladelab(node = 181, label = "GSH05", fontsize = 2, fontface = 1, offset = 0.01, offset.text = 0.0001) +
  # #
  
  # label clades - TEP GSHs
  geom_cladelab(node = 267, label = "GSH01",
                # textcolor = "purple3",
                barcolor = "purple3",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.005, offset.text = 0.0002) +
  geom_cladelab(node = 248, label = "GSH09b",
                # textcolor = "gold2",
                barcolor = "gold2",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.007, offset.text = 0.0002) +
  geom_cladelab(node = 227, label = "GSH09c_ETP",
                # textcolor = "darkorange",
                barcolor = "darkorange",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.0085, offset.text = 0.0002) +
  geom_cladelab(node = 158, label = "GSH13c_ETP",
                # textcolor = "turquoise2",
                barcolor = "turquoise2",
                barsize = 0.8,
                fontsize = 2.5, fontface = 2, offset = 0.008, offset.text = 0.0002) +
  
  # label clades - TEP nominal spp.
  geom_cladelab(node = 267, label = "P. effusa", fontsize = 3, fontface = 3, offset = 0.01, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 248, label = "P. meandrina", fontsize = 3, fontface = 3, offset = 0.013, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 227, label = "P. capitata", fontsize = 3, fontface = 3, offset = 0.0155, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 158, label = "P. lacera", fontsize = 3, fontface = 3, offset = 0.016, offset.text = 0.0004, barcolor = NA) +
  # label clades - absent nominal spp.
  geom_cladelab(node = 207, label = "P. damicornis", fontsize = 3, fontface = 3, offset = 0.012, offset.text = 0.0004, barcolor = NA) +
  geom_cladelab(node = 181, label = "P. acuta", fontsize = 3, fontface = 3, offset = 0.016, offset.text = 0.0004, barcolor = NA) +
  
  

  # setup tree visualization
  geom_rootedge(rootedge = 0.001) +
  geom_treescale(width = 0.01, x = 0, y = -1, linesize = 0.75) +
  ggplot2::xlim(-0.001, 0.05) +
  scale_color_manual(values = c("red", "navyblue", "black")) +
  scale_size_manual(values = c(1.85, 1.85, 1.5, 1.5)) +
  theme(plot.margin = margin(2,2,2,2, unit = "mm"),
        legend.position = "none",
        plot.title = element_blank(),
        plot.subtitle = element_text(size = 10))

#
pdf("./outputs/figures/FigS13_iqtree_rep_plus3b_75pct_tree_annotated_revised.pdf", width = 6.5, height = 7)
print(p_annotated)
dev.off()
```


### All no-clones IQ-TREE
```{r}
tree <- read.nexus("./outputs/iqtree/iqtree_phyluce_all_noclones.treefile")


tree <- treeio::read.iqtree("./outputs/iqtree/iqtree_phyluce_all_noclones.treefile")

tree <- read.newick("./outputs/iqtree/iqtree_phyluce_spptest_clean.treefile")
# midpoint root
tree <- midpoint.root(tree)
# set root-edge
tree$root.edge <- 1
gg_tree(tree)
```
```{r modify_tip_labels}
# remove contigs suffix 
tips <- gsub("_S.*_contigs$", "", tree$tip.label)
tree$tip.label <- tips
# remove biorepository codes 
tips_short <- gsub("_AP.*", "", tips)
tree$tip.label <- tips_short
```

```{r basic_ggtree_viz}
pdf("./outputs/figures/iqtree_03.pdf", width = 8.5, height = 11)
tree_title <- expression(paste("IQ-TREE Maximum Likelihood Tree of ", italic("Pocillopora"), " samples"))
loci_info <- c("1,986 UCE loci, 714,025 bp, 67,422 informative sites")
#
tree %>% ggtree(., right = TRUE) %<+% samples + 
  geom_tippoint(aes(fill = `Species`), shape = 21, alpha = 0.8, stroke = 0.5, color = "black", hjust = 0.3, show.legend = TRUE) +
  geom_tiplab(size = 1.5, hjust = -0.08) +
  ggplot2::xlim(-0.05, 0.25) +
  geom_treescale(width = 0.04, x = -0.025, y = 1) +
  scale_fill_manual(values = spp_colors, name = "Species") +
   theme(legend.position = c(0.12, 0.88),
        legend.margin = margin(0,0,0,0, unit = "mm"),
        plot.margin = margin(5,5,5,5, unit = "mm")) +
  ggtitle(tree_title, subtitle = loci_info)
dev.off()
```
```{r annotated_ggtree_viz}
# clades - node 103: P. effusa, node 108: P. grandis, node 180: P. meandrina, node 183: P. verrucosa-like
pdf("./outputs/figures/iqtree_02.pdf", width = 8.5, height = 11.5)
#
tree %>% ggtree(., right = TRUE) %<+% samples + 
  geom_tippoint(aes(fill = `mtorf_type`), shape = 21, alpha = 0.8, stroke = 0.5, color = "black", hjust = 0.3, show.legend = TRUE) +
  geom_tiplab(size = 3, hjust = -0.05) +
  # add bootstrap support values
  geom_nodelab(size = 2, hjust = 1.25, vjust = -0.25) +
  # add clade annotations
  geom_cladelab(node = 103, label = "P. effusa (Clipperton)", angle = 0, 
                  fontsize = 3, offset = 0.15) +
  geom_cladelab(node = 108, label = "P. grandis-like", angle = 0, 
                  fontsize = 3, offset = 0.15) +
  geom_cladelab(node = 180, label = "P. meandrina-like", angle = 0, 
                  fontsize = 3, offset = 0.15) +
  geom_cladelab(node = 183, label = "P. verrucosa-like", angle = 0, 
                  fontsize = 3, offset = 0.15) +
  ggplot2::xlim(-0.05, 0.5) +
  geom_treescale(width = 0.05, x = 0) +
  scale_fill_manual(values = mtorf_colors, name = "mtORF Type") +
   theme(legend.position = "right",
        legend.margin = margin(0,0,0,0, unit = "mm"),
        plot.margin = margin(5,5,5,5, unit = "mm")) +
  ggtitle(tree_title)
dev.off()
```
```{r}
pdf("./outputs/figures/iqtree_nodes.pdf", width = 11.5, height = 8)
tree %>% ggtree() + geom_text(aes(label=node), size = 2, hjust=.45)
dev.off()
```

#### Spp. Subset Samples Tree
```{r}
tree <- read.newick("./outputs/iqtree/iqtree_phyluce_sppsubset.treefile")
# midpoint root
tree <- midpoint.root(tree)
# set root-edge
tree$root.edge <- 1
gg_tree(tree)
```
```{r}
ggtree(tree, ladderize = T)
```

#### Spp. Subset Samples Plus Ref. Genomes Tree
```{r}

```

#### All No-Clones Samples Tree
```{r}
tree <- read.nexus("./outputs/iqtree/iqtree_phyluce_all_noclones.treefile")
# midpoint root
tree <- midpoint.root(tree)
# set root-edge
tree$root.edge <- 1
gg_tree(tree)
```
```{r modify_tip_labels}
# remove contigs suffix 
tips <- gsub("_S.*_contigs$", "", tree$tip.label)
tree$tip.label <- tips
```
```{r}
tree_title <- expression(paste("IQ-TREE Maximum Likelihood Tree of ", italic("Pocillopora"), " samples"))
#
loci_info <- c("619 UCE loci, 241,332 sites, 21,325 informative sites")
#
tree %>% ggtree(., right = TRUE) %<+% samples_k6 + 
  geom_tippoint(aes(fill = `NGSAdmix K6 Species`), shape = 21, alpha = 0.8, stroke = 0.5, color = "black", hjust = 0.3, show.legend = TRUE) +
  geom_tiplab(size = 1.5, hjust = -0.08) +
  ggplot2::xlim(-0.05, 0.25) +
  geom_treescale(width = 0.04, x = -0.025, y = 1) +
  scale_fill_manual(values = ngsadmix_pop_colors_k6, name = "Species") +
   theme(legend.position = c(0.12, 0.88),
        legend.margin = margin(0,0,0,0, unit = "mm"),
        plot.margin = margin(5,5,5,5, unit = "mm")) +
  ggtitle(tree_title, subtitle = loci_info)
```


#### All No-Clones Samples Plus Ref. Genomes Tree
```{r}

```




