---
title: "PCAngsd"
author: "Mike Connelly"
date: "2023-05-10"
output: html_document
---
## Setup and packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
options(stringsAsFactors = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```
```{r packages}
library("tidyverse")
#
source("./R/pocillopora_etp_gksim_functions.R")
```

## Colors and theme
```{r colors}
region_colors <- c("gold1", "steelblue", "darkorange", "darkturquoise")
# 
site_colors <- c("gold1", "blue", "darkblue", "steelblue", "lightblue", "darkorange3","darkorange", "orange", "orangered", "cyan", "cyan3", "cyan4", "turquoise", "lightgreen")
names(site_colors)
# 
theme_set(theme_bw())
```

## Import sample metadata
```{r import_sample_metadata}
all_samples <- read_csv("./data/pocillopora_samples_all_metadata.csv")
etp_samples <- read_csv("./data/etp_samples.txt")
```
```{r select_samples}
samples <- filter(all_samples, all_samples$`Sample ID` %in% etp_samples$`Sample ID`)
samples$`Region:Site` <- paste(samples$Region, samples$Site, sep = ":")
```
Note: Remove clones from the dataset using IBS analysis before running PCAngsd

## Perform PCA of all samples
```{r pca}
# pop<-read.table("Demo/Data/Demo1pop.info")
C <- as.matrix(read.table("./outputs/angsd/etp.cov"))
e <- eigen(C)
```
```{r pca_test_plot}
plot(e$vectors[,1:2],xlab="PC1",ylab="PC2", main="individual allele frequency")
#
plot(e$vectors[,2:3],xlab="PC2",ylab="PC3", main="individual allele frequency")
# 
```
```{r combine_pca_metadata}
# combine pc coordinates with sample metadata
pcs <- as.data.frame(e$vectors) %>% cbind(samples)
# store variable with % variance explained by each PC
pcvar <- round(e$values, 2)
#
```
### Plot PCA of all-samples
```{r pca_test_ggplot}
title <- expression(paste("Preliminary PCA of 122 ETP ", italic("Pocillopora"), " samples"))
# test ggplot 
pcs %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Region`), shape = 21, size = 2, alpha = 0.75) +
  scale_fill_manual(values = region_colors) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar[2]/pcvar[1]) + 
  xlab(paste0( "PC1 (", pcvar[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar[2], "%)")) +
  ggtitle(title)
# 
pcs %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `mtorf_type`), shape = 21, size = 2, alpha = 0.75) +
  # scale_fill_manual(values = region_colors) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar[2]/pcvar[1]) +
  xlab(paste0( "PC1 (", pcvar[1], "%)")) +
  ylab(paste0( "PC2 (", pcvar[2], "%)")) +
  ggtitle(title)
# 
pcs %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Region`, shape = `mtorf_type`), size = 2, alpha = 0.75) +
  scale_fill_manual(values = region_colors) +
  scale_shape_manual(values = c(21:25)) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar[2]/pcvar[1]) + 
  xlab(paste0( "PC1 (", pcvar[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar[2], "%)")) +
  ggtitle(title)
# 
```

```{r figure 1: pca_1-2}
# use custom function to plot PCA colored by Region:Site
pdf("./outputs/figures/pca12.pdf", width = 8, height = 6)
gg_pcangsd(e, pc=c(1,2)) +
  ggtitle(title)
  #
gg_pcangsd(e, pc=c(1,2)) +
  facet_grid(Region ~ .) +
  ggtitle(title)
dev.off()
```
```{r pca_2-3}
pdf("./outputs/figures/pca23.pdf", width = 8, height = 6)
gg_pcangsd(e, pc=c(2,3)) +
  ggtitle(title)
  #
dev.off()
```
### Assign samples to PCA clusters - 
# REDO BASED ON MTORF RESULTS!

Based on the PCA plot of all eastern Pacific *Pocillopora* samples, there appear to be three genetic clusters that separate along PC1: cluster 1 (PC1 > 0.025), cluster 2 (PC1 < 0.025 and PC1 > -0.10), and cluster 3 (PC1 < -0.10). Samples were labelled according to their genetic cluster, and PCA was performed again.

```{r assign_pca_clusters}
# assign PCA clusters based on separation along PC1
pcs$cluster <- ifelse(pcs$V1 > 0.025, "1", ifelse(pcs$V1 < 0.025 & pcs$V1 > -0.1, "2", "3"))
# extract cluster identifications
samples_pc_clusters <- pcs %>% select(`Sample ID`:`cluster`)
# 
samples_cluster1 <- samples_pc_clusters %>% filter(`cluster` == 1)
samples_cluster2 <- samples_pc_clusters %>% filter(`cluster` == 2)
samples_cluster3 <- samples_pc_clusters %>% filter(`cluster` == 3)
```
```{r pca_cluster_ggplot}
# test ggplot 
pcs %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `cluster`), shape = 21, size = 2, alpha = 0.75) +
  # scale_fill_manual(values = region_colors) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar[2]/pcvar[1]) + 
  xlab(paste0( "PC1 (", pcvar[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar[2], "%)")) +
  ggtitle(title)
# 
```

```{r write_cluster_assignments}
# write files containing a list of Sample IDs for each cluster to use for ANGSD and PCAngsd
write(samples_cluster1$`Sample ID`, file = "./data/etp_cluster1_samples.txt")
write(samples_cluster2$`Sample ID`, file = "./data/etp_cluster2_samples.txt")
write(samples_cluster3$`Sample ID`, file = "./data/etp_cluster3_samples.txt")
```


## Perform separate PCAs for different species

### Plot PCAs for *P. grandis*
```{r pca_c1}
pgrandis <- read_csv("./data/pgrandis_samples.txt", col_names = "Sample ID") %>% arrange(`Sample ID`)
pgrandis_samples <- pgrandis %>% left_join(samples)
# pop<-read.table("Demo/Data/Demo1pop.info")
C1 <- as.matrix(read.table("./outputs/angsd/pgrandis.cov"))
e1 <- eigen(C1)
```
```{r pca_test_plot}
plot(e1$vectors[,1:2],xlab="PC1",ylab="PC2", main="individual allele frequency")
#
plot(e1$vectors[,2:3],xlab="PC2",ylab="PC3", main="individual allele frequency")
# 
```
```{r combine_pca_metadata}
# combine pc coordinates with sample metadata
pcs_c1 <- as.data.frame(e1$vectors) %>% cbind(pgrandis_samples)
# store variable with % variance explained by each PC
pcvar_c1 <- round(e1$values, 2)
#
```
```{r pca_ggplot}
# test ggplot 
pcs_c1 %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Region`), shape = 21, size = 2, alpha = 0.75) +
  scale_fill_manual(values = region_colors) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar_c1[2]/pcvar_c1[1]) + 
  theme(legend.position = "right") +
  xlab(paste0( "PC1 (", pcvar_c1[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar_c1[2], "%)"))
# 
```
```{r pgrandis_region_pca}
pdf("./outputs/figures/pgrandis_region_pca.pdf", height = 4.5, width = 6)
title <- expression(paste("PCA of ETP ", italic("P. grandis"), " samples (n = 73)"))
#
pcs_c1 %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Region`), shape = 21, size = 2, alpha = 0.75) +
  scale_fill_manual(values = region_colors) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar_c1[2]/pcvar_c1[1]) + 
  theme(legend.position = "right") +
  xlab(paste0( "PC1 (", pcvar_c1[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar_c1[2], "%)")) +
  ggtitle(title)
# 
dev.off()
```


### Plot PCAs for Cluster 2
```{r}

```

### Plot PCAs for Cluster 3
```{r pca_c3}
# pop<-read.table("Demo/Data/Demo1pop.info")
C3 <- as.matrix(read.table("./outputs/angsd/etp_cluster3.cov"))
e3 <- eigen(C3)
```
```{r pca_test_plot}
plot(e$vectors[,1:2],xlab="PC1",ylab="PC2", main="individual allele frequency")
#
plot(e$vectors[,2:3],xlab="PC2",ylab="PC3", main="individual allele frequency")
# 
```
```{r combine_pca_metadata}
# combine pc coordinates with sample metadata
pcs_c3 <- as.data.frame(e3$vectors) %>% cbind(samples_cluster3)
# store variable with % variance explained by each PC
pcvar_c3 <- round(e3$values, 2)
#
```
```{r pca_cluster_ggplot}
# test ggplot 
pcs_c3 %>% ggplot(aes(x = V1, y = V2)) +
  geom_point(aes(fill = `Region`), shape = 21, size = 2, alpha = 0.75) +
  scale_fill_manual(values = region_colors) +
  # facet_grid(Region ~ .) +
  coord_fixed(pcvar_c3[2]/pcvar_c3[1]) + 
  xlab(paste0( "PC1 (", pcvar_c3[1], "%)")) + 
  ylab(paste0( "PC2 (", pcvar_c3[2], "%)"))
# 
```

