---
title: "symABCD"
author: "Mike Connelly"
date: "4/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(stringsAsFactors = FALSE)
```

```{r}
library("tidyverse") # for data wrangling and plotting
library("magrittr")
library("RColorBrewer") # for colors
```
## Import sample metadata
```{r import_sample_metadata}
source("./R/sample_import.R")
samples <- all_samples
```
```{r select_samples}
samples <- filter(all_samples, ANALYSIS ==T)
# samples <- filter(all_samples, `PANAMA_MANUSCRIPT` == T)
```

## Plot the proportion of reads mapping to A, B, C, and D in Pocillopora corals
```{r}
abcd_pdam <- read.delim("./outputs/symbiont_counts/allcounts_pgra_himb_symCD.txt")

df_abcd_pdam <- abcd_pdam %>% 
  dplyr::select(sample, fracA, fracB, fracC, fracD) %>% drop_na() %>% 
  # reorder samples based on proportion of Cladocopium
  mutate("sample" = forcats::fct_reorder2(`sample`, desc(fracC), desc(fracD))) %>% 
  gather(clade, frac, fracA:fracD) %>% 
  dplyr::mutate(Genus = factor(ifelse(clade == "fracA", "Symbiodinium", ifelse(clade == "fracB", "Breviolum", ifelse(clade == "fracC", "Cladocopium", "Durusdinium"))))) %>%
  dplyr::mutate(`Sample ID` = str_extract(sample, "^.*_.*_[S]*")) %>% 
  dplyr::mutate(`Sample ID` = str_replace(`Sample ID`, "_S.*_", "")) %>% 
  # dplyr::filter(`Sample ID` != "Perlas_344_AP3XG35") #%>% 
  dplyr::left_join(samples, by = "Sample ID")
# 
df_abcd_pdam$Genus <- factor(df_abcd_pdam$Genus, levels = c("Symbiodinium", "Breviolum", "Cladocopium", "Durusdinium", ordered = T))

plot_abcd_pdam <- df_abcd_pdam %>% 
  ggplot() + 
  geom_bar(aes(y = frac, x = sample, fill = Genus), stat="identity") +
  facet_grid(~Region, scales = "free") +
  # facet_grid(Region ~ cluster, scales = "free") +
  scale_fill_brewer(palette="Spectral", name = "Genus")+
  theme_classic()+
  ylab("Proportion of Symbiont Reads")+
  xlab("Pocillopora Samples")+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```
```{r}
pdf("./outputs/figures/pocillopora_symbiont_proportions_species_region_facet.pdf", width = 8, height = 8.75)
print(plot_abcd_pdam)
dev.off()
```
### Exploratory figures
```{r }
df_abcd_pdam %<>% left_join(samples, by = "Sample ID") 
###

df_abcd_pdam %<>% filter(NOCLONES_SAMPLE)
df_abcd_pdam %<>% filter(Species == "P. grandis")

df_abcd_pdam %<>% filter(Set == "USFQ_CSUMB")
# df_abcd_pdam$Genotype
# 
plot_abcd_pdam <- df_abcd_pdam %>% 
  ggplot() + 
  geom_bar(aes(y = frac, x = sample, fill = Genus), stat="identity")+
  facet_wrap(~ `Region:Site`, scales = "free") +
  scale_fill_brewer(palette="Spectral", name = "Genus")+
  theme_classic()+
  ylab("Proportion of Symbiont Reads")+
  xlab(NULL)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

plot_abcd_pdam <- df_abcd_pdam %>% 
  ggplot() + 
  geom_bar(aes(y = frac, x = sample, fill = Genus), stat="identity")+
  facet_grid(`Species` ~ ., scales = "free", space = "free") +
  scale_fill_brewer(palette="Spectral", name = "Genus")+
  theme_classic()+
  ylab("Proportion of Symbiont Reads")+
  xlab(NULL)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

plot_abcd_pdam <- df_abcd_pdam %>% 
  ggplot() + 
  geom_bar(aes(y = frac, x = sample, fill = Genus), stat="identity")+
  facet_wrap(`Region` + `Site` ~ ., scales = "free") +
  scale_fill_brewer(palette="Spectral", name = "Genus")+
  theme_classic()+
  ylab("Proportion of Symbiont Reads")+
  xlab(NULL)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "right")

```
```{r}
print(plot_abcd_pdam)
```


### Figure S13
```{r}
# THIS ONE
plot_abcd_pdam <- df_abcd_pdam %>% 
  ggplot() + 
  geom_bar(aes(y = frac, x = sample, fill = Genus), stat="identity")+
  facet_wrap(`Species` ~ ., scales = "free") +
  scale_fill_brewer(palette="Spectral", name = "Genus")+
  theme_classic()+
  ylab("Proportion of Symbiont Reads")+
  xlab(NULL)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = c(0.85, 0.25)) +
  theme(axis.title = element_text(size = 10),
        axis.text = element_text(size = 8),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 10))

print(plot_abcd_pdam)
```
```{r}
pdf("./outputs/figures/FigS13_CD_proportions_species.pdf", width = 6.5, height = 5.5)
print(plot_abcd_pdam)
dev.off()
```


### Symbiont proportion by species and region
```{r}
plot_abcd_pdam <- df_abcd_pdam %>% 
  ggplot() + 
  geom_bar(aes(y = frac, x = sample, fill = Genus), stat="identity")+
  facet_wrap(`Species` + `Site` ~ ., scales = "free") +
  scale_fill_brewer(palette="Spectral", name = "Genus")+
  theme_classic()+
  ylab("Proportion of Symbiont Reads")+
  xlab(NULL)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = c(0.85, 0.1)) +
  theme(plot.title = element_text(size = 16, hjust = -0.2),
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)) +
  ggtitle(expression(paste("East Pacific ", italic("Pocillopora"), " species symbiont community variation")))
```
```{r}
print(plot_abcd_pdam)
```

## Scatterpie plot on map 
to examine geographic patterns
```{r}
all(df_abcd_pdam$`Sample ID` %in% samples$`Sample ID`)
```





## Pie chart tips on IBS clustering tree
to examine phylogenetic and within-clone patterns
```{r}
df_plot <- df_abcd_pdam %>% dplyr::select(`Sample ID`, clade:Genus)
df_plot_C <- df_plot %>% filter(clade=="fracC")
```
```{r}
tr <- rtree(30)
tr <- tree

d1 <- data.frame(id=tr$tip.label, val=rnorm(99, sd=3))
p <- ggtree(tr)

p2 <- facet_plot(p, panel="dot", data=d1, geom=geom_point, aes(x=val), color='firebrick')
#
d2 <- data.frame(id=tr$tip.label, value = abs(rnorm(99, mean=100, sd=50)))

facet_plot(p2, panel = 'bar', data = d2, geom=geom_segment, aes(x=0, xend=value, y=y, yend=y), size=3, color='steelblue') + theme_tree2()

facet_plot(p, panel = 'symbiont proportion', data = df_plot_C, geom = geom_point, mapping = aes(x = frac, fill = Genus), size=1) + theme_tree2()

df_plot_C$`Sample ID` %in% tree$tip.label
```
```{r}
ggplot(df_abcd_pdam) + geom_col(aes(x = frac, y = sample, fill = Genus))
```

```{r}
library(ggtree)
library(reshape2)
library(ggplot2)

set.seed(2020-12-18)

## generate random data with 15 tips
num_tips = 15
tr <- rtree(num_tips)
p <- ggtree(tr)

a <- runif(num_tips, 0, 0.33)
b <- runif(num_tips, 0, 0.33)
c <- runif(num_tips, 0, 0.33)
d <- 1 - a - b - c
dat <- data.frame(a=a, b=b, c=c, d=d)

## melt the dataframe to create pie-charts
dat.m = melt(dat, id.vars = "tip")
dat.m = melt(dat, id.vars = "tip")

## create list to hold all pie charts
pies = list()
for (i in 1:num_tips) {
  curr_dat = melt(dat[i,])
  ## create a ggplot object for each pie chart
  pies[[i]] =  ggplot(curr_dat, aes(y = value, fill = variable, x="")) + 
    geom_bar(stat = "identity") +
    coord_polar("y", start=0) +
    theme_void() + scale_fill_brewer(palette = "Set1", guide = F)
}
# give them the appropriate names and plot on tree
names(pies) = 1:15
inset(p, pies, width=0.1, height=0.1, hjust=0)
```

## Categorize samples as C-dominated, D-dominated, or mixed
```{r}
pdam_cd_cats <- abcd_pdam %>%
  dplyr::mutate(`Sample ID` = str_extract(sample, "^.*_.*_[S]*")) %>% 
  dplyr::mutate(`Sample ID` = str_replace(`Sample ID`, "_S.*_", "")) %>% 
  # dplyr::filter(`Sample ID` != "Perlas_344_AP3XG35") #%>% 
  dplyr::left_join(samples, by = "Sample ID") %>% 
  filter(NOCLONES_SAMPLE) %>% 
  dplyr::left_join(samples_k5_assignments, by = "Sample ID") %>% 
  mutate(`Symbiont Community` = ifelse(fracC > 0.9, "C-dominated", ifelse(fracD > 0.9, "D-dominated", "Mixed")))

pdam_cd_cats %>% group_by(`Symbiont Community`) %>% count()

pdam_cd_cats %>% group_by(`NGSAdmix K5 Species`, `Symbiont Community`) %>% count() %>% pivot_wider(names_from = `NGSAdmix K5 Species`, values_from = n) %>% View()

pdam_cd_cats %>% ggplot(aes(fracD, fracC)) + geom_point(aes(color = `Symbiont Community`)) + facet_wrap(~`NGSAdmix K5 Species`)
```

